---
title: "Single-Cell Cell-Cell Interaction Arrays - Migration Analysis"  
author: "Haylie Helms"
date: "2025-10-29"
output:
  pdf_document:
    latex_engine: xelatex
    keep_tex: true
  html_document: default
geometry: margin=1in
fontsize: 11pt
mainfont: Arial
monofont: Menlo
header-includes:
  - \usepackage{microtype}
  - \usepackage{xurl}
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines=true,breakanywhere=true,commandchars=\\\{\}}
  - \sloppy
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
library(readxl)
library(stringr)
library(dplyr)
library(ggplot2)
library(janitor) 
library(purrr)
library(readr)
library(ggpubr)
library(geometry)
library(tidyr)
library(scales)
library(RANN)

# Plot Formatting
celltype_colors <- c(
  "MDAMB231"     = "#B64249",
  "Myoepithelial"= "#796089",
  "Fibroblast"   = "#718F77",
  "Missing_ID"   = "#AFABAB")

group_colors <- c("Den1" = "#DD9358", "Den10" = "#6A7AA1")

point_colors <- c("Den1" = "#D6986A", "Den10" = "#6B748D")
```

# Cancer Density
## Load & preprocess data
```{r}
folder <- "/Users/helmsh/Desktop/"

files <- c(
  "Den1S1.xlsx","Den1S2.xlsx","Den1S3.xlsx",
  "Den10S1.xlsx","Den10S2.xlsx","Den10S3.xlsx")

# Extract sample names
sample_names <- files %>%
  str_remove("^Den") %>%
  str_remove("\\.xlsx$")

# Load each file, read the "Position" sheet, skip row 1 (a header), treat everything as text, rename column 7 to CellTypeID (inconsistent naming during export), and add the sample name as a new column
data_list <- lapply(seq_along(files), function(i) {
  read_excel(
    path = file.path(folder, files[i]),
    sheet = "Position",
    skip = 1,
    col_types = "text"
  ) %>%
    rename(CellTypeID = 7) %>%
    mutate(SampleID = sample_names[i])  # add the sample name as a column
})

# Combine into one dataframe
all_positions <- bind_rows(data_list)
```

```{r}
head(all_positions)
unique(all_positions$CellTypeID)
unique(all_positions$SampleID)
```

```{r}
# Issue with file export from Imaris. The annotations have different names. Create function to deal with that

all_positions <- all_positions %>%
  mutate(
    CellTypeID = case_when(
      grepl("^MDA", CellTypeID, ignore.case = TRUE) ~ "MDAMB231",
      grepl("^Myo", CellTypeID, ignore.case = TRUE) ~ "Myoepithelial",
      is.na(CellTypeID) ~ "Missing_ID",
      TRUE ~ CellTypeID))

unique(all_positions$CellTypeID)
```

```{r}
## Add Unique ID - Imaris reuses track ID's across files
all_positions <- all_positions %>%
  mutate(UniqueID = paste(SampleID, TrackID, sep = "_"))

head(all_positions)
```

```{r}
# Confirm we got all of them. These datasets have 145 frames. So the max shown should be 145 instances. If there are more that means there is a duplicate still
all_positions %>% 
  count(UniqueID) %>% 
  arrange(desc(n))
```

```{r}
## Inspecting missing track annotations
missing_tracks <- all_positions %>%
  filter(CellTypeID == "Missing_ID") %>%
  distinct(SampleID, UniqueID)

missing_tracks
n_missing <- nrow(missing_tracks)
n_missing
```

```{r}
# Remove spaces in col names, i.e. "Position X" to "X_Position". Spaces cause problems. Also renaming ID because it seemed to want to pull functions instead of data
all_positions <- all_positions %>%
  rename(
    X_Position = `Position X`,
    Y_Position = `Position Y`,
    SpotID     = `ID`)

head(all_positions)
```

```{r}
# Make SpotID all the same number of characters
all_positions$SpotID <- str_pad(all_positions$SpotID, width = 6, side = "left", pad = "0")

head(all_positions)
```

```{r}
# One of the short track ID's is 000000. Rename to 999999
all_positions$SpotID[all_positions$SpotID == "000000"] <- "999999"

head(all_positions)
```

```{r}
# Make time all the same number of characters
all_positions$Time <- str_pad(all_positions$Time, width = 3, side = "left", pad = "0")

head(all_positions)
```

```{r}
# Reorder so it is easier to work with
all_positions_cleaned <- all_positions %>%
  select(SampleID, UniqueID, CellTypeID, Time, X_Position, Y_Position, Unit, Category, Collection, TrackID, SpotID)

head(all_positions_cleaned)
```

```{r}
# Rearrange so that is sorted by SampleID, then UniqueID, then Time so we can separate out the duplicates
all_positions_cleaned <- all_positions_cleaned %>% arrange(SampleID, UniqueID, Time)

head(all_positions_cleaned)
```


```{r}
## Add in Group - Den1 vs Den10
Combined_Positions <- all_positions_cleaned %>%
  mutate(
    Group = case_when(
      str_starts(SampleID, "1S")  ~ "Den1",
      str_starts(SampleID, "10S") ~ "Den10",
      TRUE ~ NA_character_ ))

head(Combined_Positions)
```

## Analysis
### 1.) Migration Tracks
##### Every Sample Individually, Each Channel, and Merged
```{r}
Combined_Positions <- Combined_Positions %>%
  mutate(
    TimeNum = as.integer(Time),
    X = as.numeric(X_Position),
    Y = as.numeric(Y_Position)
  ) %>%
  arrange(SampleID, UniqueID, TimeNum)

# helper for clean numeric labels
.lab_int <- function(x) format(round(x), trim = TRUE, scientific = FALSE,
                               big.mark = "", decimal.mark = ".")

make_sample_plot <- function(df, sample_name, title_text,
                             span_x = 800, span_y = 800, tick = 200) {

  lims <- dplyr::filter(sample_limits, SampleID == sample_name)
  stopifnot(nrow(lims) == 1)

  # default: centered 800×800
  xlim <- c(lims$xmid - span_x/2, lims$xmid + span_x/2)
  ylim <- c(lims$ymid - span_y/2, lims$ymid + span_y/2)

  # shift 1S1 vertically so ymin touches x-axis
  if (identical(sample_name, "1S1")) {
    ylim <- c(lims$ymin0, lims$ymin0 + span_y)
  }

  x_breaks <- seq(xlim[1], xlim[2], by = tick)
  y_breaks <- seq(ylim[1], ylim[2], by = tick)

  ggplot(df, aes(x = X, y = Y, group = UniqueID, color = CellTypeID)) +
    geom_path(linewidth = 0.6, alpha = 0.6, na.rm = TRUE) +
    coord_equal() +
    scale_color_manual(values = celltype_colors) +
    scale_x_continuous(limits = xlim, breaks = x_breaks, labels = .lab_int,
                       expand = c(0, 0), name = expression(X~(mu*m))) +
    scale_y_continuous(limits = ylim, breaks = y_breaks, labels = .lab_int,
                       expand = c(0, 0), name = expression(Y~(mu*m))) +
    theme_classic(base_size = 20) +
    theme(legend.position = "none",
          axis.title = element_text(size = 20),
          axis.text  = element_text(size = 20),
          plot.title = element_text(size = 20, face = "bold")) +
    labs(title = title_text)
}


# Loop & save
sample_limits <- Combined_Positions %>%
  group_by(SampleID) %>%
  summarise(
    xmin  = min(X, na.rm = TRUE),
    xmax  = max(X, na.rm = TRUE),
    ymin  = min(Y, na.rm = TRUE),
    ymax  = max(Y, na.rm = TRUE),
    xmid  = (xmin + xmax) / 2,
    ymid  = (ymin + ymax) / 2,
    ymin0 = ymin,
    .groups = "drop")

unique_samples <- unique(Combined_Positions$SampleID)

for (s in unique_samples) {
  df_s <- dplyr::filter(Combined_Positions, SampleID == s)

  p_all <- make_sample_plot(df_s, s, paste0(s, " — All Tracks (Raw)"))
  ggsave(paste0(s, "_Raw_All_Tracks.png"), p_all, width = 6, height = 4, dpi = 600)

  p_mda <- make_sample_plot(dplyr::filter(df_s, CellTypeID == "MDAMB231"),
                            s, paste0(s, " — MDAMB231 Tracks (Raw)"))
  ggsave(paste0(s, "_Raw_MDAMB231_Tracks.png"), p_mda, width = 6, height = 4, dpi = 600)

  p_myo <- make_sample_plot(dplyr::filter(df_s, CellTypeID == "Myoepithelial"),
                            s, paste0(s, " — Myoepithelial Tracks (Raw)"))
  ggsave(paste0(s, "_Raw_Myoepithelial_Tracks.png"), p_myo, width = 6, height = 4, dpi = 600)
}
```

#### Center Aligned
```{r}
# Translate each Sample so their bounding boxes are all centered

# Numeric coords
CP_num <- Combined_Positions %>%
  mutate(
    X = as.numeric(X_Position),
    Y = as.numeric(Y_Position),
    TimeN = as.integer(Time)
  )

# Center each SampleID directly by subtracting its bbox midpoint
aligned <- CP_num %>%
  group_by(SampleID) %>%
  mutate(
    x_mid = (min(X, na.rm = TRUE) + max(X, na.rm = TRUE)) / 2,
    y_mid = (min(Y, na.rm = TRUE) + max(Y, na.rm = TRUE)) / 2,
    x_centered = X - x_mid,
    y_centered = Y - y_mid
  ) %>%
  ungroup()

# One-off vertical shift for 1S1 so ymin == -450 on the fixed axes
if ("1S1" %in% aligned$SampleID) {
  ymin_1S1 <- aligned %>%
    dplyr::filter(SampleID == "1S1") %>%
    summarise(min_y = min(y_centered, na.rm = TRUE)) %>%
    pull(min_y)

  shift_1S1 <- (-450) - ymin_1S1  # amount to add so min becomes -450

  aligned <- aligned %>%
    mutate(y_centered = ifelse(SampleID == "1S1",
                               y_centered + shift_1S1,
                               y_centered))
}

# Sanity check of global bounds (after centering each sample)
aligned %>%
  summarise(
    xmin = min(x_centered, na.rm = TRUE),
    xmax = max(x_centered, na.rm = TRUE),
    ymin = min(y_centered, na.rm = TRUE),
    ymax = max(y_centered, na.rm = TRUE)
  )

```

##### Aligned - Cycle through all plot variations
```{r}
plot_tracks_fixed <- function(df) {
  ggplot(
    df,
    aes(
      x = x_centered,
      y = y_centered,
      group = interaction(SampleID, UniqueID),
      color = CellTypeID
    )
  ) +
    geom_path(linewidth = 0.6, alpha = 0.6, na.rm = TRUE) +
    coord_equal() +
    scale_x_continuous(
      limits = c(-445, 450),
      breaks  = c(-400, -200, 0, 200, 400),
      expand  = c(0, 0)
    ) +
    scale_y_continuous(
      limits = c(-450, 450),
      breaks  = c(-400, -200, 0, 200, 400),
      expand  = c(0, 0)
    ) +
    scale_color_manual(values = celltype_colors) +
    theme_classic(base_size = 20) +
    theme(
      legend.position = "none",         
      plot.title      = element_blank(), 
      axis.title      = element_text(size = 20),
      axis.text       = element_text(size = 20),
      plot.margin     = unit(c(8, 8, 8, 8), "pt")
    ) +
    labs(
      x = "X (µm)",
      y = "Y (µm)"
    )
}

den1  <- aligned %>% filter(Group == "Den1")
den10 <- aligned %>% filter(Group == "Den10")

# Den1
p_den1_all <- plot_tracks_fixed(den1)
p_den1_mda <- plot_tracks_fixed(den1 %>% filter(CellTypeID == "MDAMB231"))
p_den1_myo <- plot_tracks_fixed(den1 %>% filter(CellTypeID == "Myoepithelial"))

ggsave("Den1_Aligned_All_Tracks.png",           p_den1_all, width = 7, height = 6, dpi = 600)
ggsave("Den1_Aligned_MDAMB231_Tracks.png",      p_den1_mda, width = 7, height = 6, dpi = 600)
ggsave("Den1_Aligned_Myoepithelial_Tracks.png", p_den1_myo, width = 7, height = 6, dpi = 600)

# Den10
p_den10_all <- plot_tracks_fixed(den10)
p_den10_mda <- plot_tracks_fixed(den10 %>% filter(CellTypeID == "MDAMB231"))
p_den10_myo <- plot_tracks_fixed(den10 %>% filter(CellTypeID == "Myoepithelial"))

ggsave("Den10_Aligned_All_Tracks.png",           p_den10_all, width = 7, height = 6, dpi = 600)
ggsave("Den10_Aligned_MDAMB231_Tracks.png",      p_den10_mda, width = 7, height = 6, dpi = 600)
ggsave("Den10_Aligned_Myoepithelial_Tracks.png", p_den10_myo, width = 7, height = 6, dpi = 600)

# Per sample exports
samples <- sort(unique(aligned$SampleID))

for (s in samples) {
  d <- aligned %>% filter(SampleID == s)

  p_all <- plot_tracks_fixed(d)
  p_mda <- plot_tracks_fixed(d %>% filter(CellTypeID == "MDAMB231"))
  p_myo <- plot_tracks_fixed(d %>% filter(CellTypeID == "Myoepithelial"))

  ggsave(paste0(s, "_Aligned_All_Tracks.png"),           p_all, width = 6, height = 6, dpi = 600)
  ggsave(paste0(s, "_Aligned_MDAMB231_Tracks.png"),      p_mda, width = 6, height = 6, dpi = 600)
  ggsave(paste0(s, "_Aligned_Myoepithelial_Tracks.png"), p_myo, width = 6, height = 6, dpi = 600)
}
```

#### Relative Position
```{r}
Combined_Positions_Relative <- Combined_Positions %>%
  mutate(
    Xnum = as.numeric(X_Position),
    Ynum = as.numeric(Y_Position),
    Tnum = as.integer(Time)
  ) %>%
  arrange(SampleID, UniqueID, Tnum) %>%
  group_by(SampleID, UniqueID) %>%
  # anchor each track at its first *observed* (non-NA) position
  mutate(
    .x0 = {
      ix <- which(!is.na(Xnum))
      if (length(ix)) Xnum[ix[1]] else NA_real_
    },
    .y0 = {
      iy <- which(!is.na(Ynum))
      if (length(iy)) Ynum[iy[1]] else NA_real_
    },
    RelX = Xnum - .x0,
    RelY = Ynum - .y0
  ) %>%
  ungroup() %>%
  select(
    # keep your columns in order
    SampleID, UniqueID, CellTypeID, Time,
    X_Position, Y_Position, Unit, Category, Collection, SpotID,
    dplyr::any_of("Group"),
    RelX, RelY
  )

# Sanity check: first point of every track should be (0,0)
Combined_Positions_Relative %>%
  group_by(SampleID, UniqueID) %>%
  slice(1) %>%
  summarise(first_RelX = first(RelX), first_RelY = first(RelY), .groups = "drop") %>%
  summarise(
    n_tracks = n(),
    n_zero_start = sum(first_RelX == 0 & first_RelY == 0, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  print()
```

```{r}
# Figure out min max for myoeps to set the axis properly
myoep_pos <- Combined_Positions_Relative %>%
  dplyr::filter(CellTypeID == "Myoepithelial")

myoep_maxvals <- myoep_pos %>%
  dplyr::summarise(
    min_RelX = min(RelX, na.rm = TRUE),
    max_RelX = max(RelX, na.rm = TRUE),
    min_RelY = min(RelY, na.rm = TRUE),
    max_RelY = max(RelY, na.rm = TRUE)
  )

myoep_maxvals
```

```{r}
# Cycle through all the combos - relative position 
CP_rel <- Combined_Positions_Relative %>%
  dplyr::mutate(
    RelX   = as.numeric(RelX),
    RelY   = as.numeric(RelY)
  )

plot_tracks_rel_fixed <- function(df, title, show_legend = TRUE) {
  ggplot(
    df,
    aes(
      x = RelX,
      y = RelY,
      group = interaction(SampleID, UniqueID),
      color = CellTypeID
    )
  ) +
    geom_path(linewidth = 0.6, alpha = 0.6, na.rm = TRUE) +
    guides(color = guide_legend(override.aes = list(linewidth = 2))) +
    coord_equal() +
    scale_x_continuous(
      limits = c(-600, 600),
      breaks  = c(-600, -300, 0, 300, 600),
      expand  = c(0, 0)
    ) +
    scale_y_continuous(
      limits = c(-600, 600),
      breaks  = c(-600, -300, 0, 300, 600),
      expand  = c(0, 0)
    ) +
    scale_color_manual(values = celltype_colors) +
    theme_classic(base_size = 22) +
    theme(
      axis.line    = element_line(linewidth = 0.8, color = "black"),
      axis.text.y  = element_text(size = 14),
      axis.title.y = element_text(size = 16),
      axis.text.x  = element_text(size = 14, hjust = 0.5),
      axis.title.x = element_text(size = 16),
      plot.title   = element_text(size = 16, hjust = 0.5),
      legend.title      = element_text(size = 16),
      legend.text       = element_text(size = 14),
      legend.key.height = unit(30, "pt"),
      legend.key.width  = unit(30, "pt"),
      legend.position   = if (show_legend) "right" else "none"
    ) +
    labs(
      title = title,
      x = "Relative X Position (µm)",
      y = "Relative Y Position (µm)",
      color = "Track Identity"
    )
}

den1_rel  <- CP_rel %>% dplyr::filter(Group == "Den1")
den10_rel <- CP_rel %>% dplyr::filter(Group == "Den10")

# Den1
p_den1_all_rel <- plot_tracks_rel_fixed(den1_rel,  "Den1 — Relative: All Tracks", show_legend = TRUE)
p_den1_mda_rel <- plot_tracks_rel_fixed(
  den1_rel %>% dplyr::filter(CellTypeID == "MDAMB231"),
  "Den1 — Relative: MDAMB231 Tracks",
  show_legend = TRUE
)
p_den1_myo_rel <- plot_tracks_rel_fixed(
  den1_rel %>% dplyr::filter(CellTypeID == "Myoepithelial"),
  "Den1 — Relative: Myoepithelial Tracks",
  show_legend = TRUE
)

ggsave("Den1_Relative_All_Tracks.png",           p_den1_all_rel, width = 7, height = 6, dpi = 600)
ggsave("Den1_Relative_MDAMB231_Tracks.png",      p_den1_mda_rel, width = 7, height = 6, dpi = 600)
ggsave("Den1_Relative_Myoepithelial_Tracks.png", p_den1_myo_rel, width = 7, height = 6, dpi = 600)

# Den10
p_den10_all_rel <- plot_tracks_rel_fixed(den10_rel, "Den10 — Relative: All Tracks", show_legend = TRUE)
p_den10_mda_rel <- plot_tracks_rel_fixed(
  den10_rel %>% dplyr::filter(CellTypeID == "MDAMB231"),
  "Den10 — Relative: MDAMB231 Tracks",
  show_legend = TRUE
)
p_den10_myo_rel <- plot_tracks_rel_fixed(
  den10_rel %>% dplyr::filter(CellTypeID == "Myoepithelial"),
  "Den10 — Relative: Myoepithelial Tracks",
  show_legend = TRUE
)

ggsave("Den10_Relative_All_Tracks.png",           p_den10_all_rel, width = 7, height = 6, dpi = 600)
ggsave("Den10_Relative_MDAMB231_Tracks.png",      p_den10_mda_rel, width = 7, height = 6, dpi = 600)
ggsave("Den10_Relative_Myoepithelial_Tracks.png", p_den10_myo_rel, width = 7, height = 6, dpi = 600)

# Per sample exports
samples_rel <- sort(unique(CP_rel$SampleID))

for (s in samples_rel) {
  d <- CP_rel %>% dplyr::filter(SampleID == s)

  p_all <- plot_tracks_rel_fixed(d,  paste0(s, " — Relative: All Tracks"),       show_legend = FALSE)
  p_mda <- plot_tracks_rel_fixed(d %>% dplyr::filter(CellTypeID == "MDAMB231"),
                                 paste0(s, " — Relative: MDAMB231 Tracks"),
                                 show_legend = FALSE)
  p_myo <- plot_tracks_rel_fixed(d %>% dplyr::filter(CellTypeID == "Myoepithelial"),
                                 paste0(s, " — Relative: Myoepithelial Tracks"),
                                 show_legend = FALSE)

  ggsave(paste0(s, "_Relative_All_Tracks.png"),           p_all, width = 6, height = 6, dpi = 600)
  ggsave(paste0(s, "_Relative_MDAMB231_Tracks.png"),      p_mda, width = 6, height = 6, dpi = 600)
  ggsave(paste0(s, "_Relative_Myoepithelial_Tracks.png"), p_myo, width = 6, height = 6, dpi = 600)
}
```






### 2.) Convex Hull - Migration territory 
```{r}
Combined_Positions_Relative <- Combined_Positions_Relative %>%
  dplyr::mutate(
    Group = dplyr::case_when(
      stringr::str_starts(SampleID, "1S")  ~ "Den1",
      stringr::str_starts(SampleID, "10S") ~ "Den10",
      TRUE ~ NA_character_
    ),
    Group = factor(Group, levels = c("Den1", "Den10"))
  )

# Per-track convex hull areas
ConvexAreas <- Combined_Positions_Relative %>%
  dplyr::group_by(SampleID, UniqueID, CellTypeID, Group) %>%
  dplyr::summarise(
    area = {
      pts <- cbind(RelX, RelY)
      pts <- pts[stats::complete.cases(pts), , drop = FALSE]
      if (nrow(pts) < 3) NA_real_ else geometry::convhulln(pts, "FA")$area
    },
    .groups = "drop"
  )

# Helpers for hull polygons
hull_df <- function(df, x = "RelX", y = "RelY") {
  pts <- df[, c(x, y)]
  pts <- pts[stats::complete.cases(pts), , drop = FALSE]
  if (nrow(pts) < 3) return(NULL)
  idx <- grDevices::chull(pts[[1]], pts[[2]])
  out <- pts[idx, , drop = FALSE]
  rbind(out, out[1, , drop = FALSE]) # close polygon
}

build_per_track_hulls <- function(dat) {
  dat %>%
    dplyr::group_by(SampleID, UniqueID, CellTypeID) %>%
    dplyr::group_modify(~{
      h <- hull_df(.x, "RelX", "RelY")
      if (is.null(h)) return(tibble::tibble(RelX = numeric(0), RelY = numeric(0)))
      tibble::tibble(RelX = h$RelX, RelY = h$RelY)
    }) %>%
    dplyr::ungroup()
}

build_sample_hull <- function(dat) {
  h <- hull_df(dat, "RelX", "RelY")
  if (is.null(h)) return(NULL)
  tibble::tibble(RelX = h$RelX, RelY = h$RelY)
}

base_style <- function(show_legend = FALSE) {
  ggplot2::theme_classic(base_size = 20) +
    ggplot2::theme(
      legend.position = if (show_legend) "right" else "none",
      axis.title = ggplot2::element_text(size = 20),
      axis.text  = ggplot2::element_text(size = 20),
      plot.title = ggplot2::element_text(size = 20, face = "bold"),
      plot.margin = ggplot2::margin(10, 10, 10, 10)
    )
}

# Plot one sample with per-track hulls + sample hull, with flexible axis breaks 
plot_sample_with_hulls <- function(sample_id,
                                   cell_filter = c("ALL", "MDAMB231", "Myoepithelial"),
                                   per_track_hulls = TRUE,
                                   sample_hull = TRUE,
                                   fixed_limits = c(-300, 300, -300, 300),
                                   axis_breaks  = c(-300, -150, 0, 150, 300),
                                   show_legend = FALSE) {

  cell_filter <- match.arg(cell_filter)

  df <- Combined_Positions_Relative %>% dplyr::filter(SampleID == sample_id)
  if (cell_filter != "ALL") df <- df %>% dplyr::filter(CellTypeID == cell_filter)
  if (!nrow(df)) return(ggplot2::ggplot() + ggplot2::ggtitle(paste0(sample_id, " — EMPTY")))

  per_track <- if (per_track_hulls) build_per_track_hulls(df) else NULL
  smp_hull  <- if (sample_hull)      build_sample_hull(df)    else NULL

  xlim <- fixed_limits[1:2]
  ylim <- fixed_limits[3:4]

  p <- ggplot2::ggplot(df, ggplot2::aes(x = RelX, y = RelY, group = UniqueID, color = CellTypeID)) +
    ggplot2::geom_path(linewidth = 0.6, alpha = 0.6, na.rm = TRUE) +
    # keep aspect ratio square and allow limiting without dropping points
    ggplot2::coord_cartesian(xlim = xlim, ylim = ylim, clip = "off") +
    ggplot2::scale_x_continuous(breaks = axis_breaks, expand = c(0, 0)) +
    ggplot2::scale_y_continuous(breaks = axis_breaks, expand = c(0, 0)) +
    ggplot2::scale_color_manual(values = celltype_colors) +
    ggplot2::scale_fill_manual(values = celltype_colors) +
    ggplot2::labs(
      title = paste0(sample_id, " — Convex Hull Area, Relative XY (", cell_filter, ")"),
      x = "Relative X Position (µm)",
      y = "Relative Y Position (µm)",
      color = "Cell Type"
    ) +
    base_style(show_legend)

  # enforce equal scaling after coord_cartesian (one coord system only)
  p <- p + ggplot2::coord_fixed(ratio = 1, xlim = xlim, ylim = ylim, clip = "off")

  if (!is.null(per_track) && nrow(per_track)) {
    per_track <- per_track %>%
      dplyr::left_join(df %>% dplyr::distinct(SampleID, UniqueID, CellTypeID),
                       by = c("SampleID", "UniqueID", "CellTypeID"))

    p <- p +
      ggplot2::geom_polygon(
        data = per_track,
        ggplot2::aes(x = RelX, y = RelY, group = UniqueID, fill = CellTypeID, color = CellTypeID),
        linewidth = 0.4, alpha = 0.15, inherit.aes = FALSE
      )
  }

  if (!is.null(smp_hull)) {
    p <- p +
      ggplot2::geom_path(
        data = smp_hull, ggplot2::aes(x = RelX, y = RelY),
        color = "black", linewidth = 1.4, inherit.aes = FALSE
      )
  }

  p
}
```

#### Hull Viz
```{r}
samples <- sort(unique(Combined_Positions_Relative$SampleID))

for (s in samples) {

  # MDAMB231
  ggplot2::ggsave(
    paste0(s, "_Rel_MDAMB231_withHulls_fixed300.png"),
    plot_sample_with_hulls(
      s, "MDAMB231",
      fixed_limits = c(-300, 300, -300, 300),
      axis_breaks  = c(-300, -150, 0, 150, 300),
      per_track_hulls = TRUE,
      sample_hull = TRUE
    ),
    width = 7, height = 6, dpi = 600
  )

  # Myoep
  ggplot2::ggsave(
    paste0(s, "_Rel_Myoepithelial_withHulls_fixed600.png"),
    plot_sample_with_hulls(
      s, "Myoepithelial",
      fixed_limits = c(-600, 600, -600, 600),
      axis_breaks  = c(-600, -300, 0, 300, 600),
      per_track_hulls = TRUE,
      sample_hull = TRUE
    ),
    width = 7, height = 6, dpi = 600
  )
}
```


##### Grouped on one plot
```{r}
build_sample_hulls <- function(dat, cell_filter = "MDAMB231") {
  df <- dat
  if (!identical(cell_filter, "ALL")) {
    df <- df %>% dplyr::filter(CellTypeID == cell_filter)
  }

  df %>%
    dplyr::group_by(Group, SampleID) %>%
    dplyr::group_modify(~{
      h <- hull_df(.x, "RelX", "RelY")
      if (is.null(h)) return(tibble::tibble(RelX = numeric(0), RelY = numeric(0)))
      tibble::tibble(RelX = h$RelX, RelY = h$RelY)
    }) %>%
    dplyr::ungroup() %>%
    dplyr::filter(!(is.na(RelX) | is.na(RelY)))
}

plot_group_hull_overlays <- function(hulls_df,
                                     group_name = c("Den1", "Den10"),
                                     fixed_limits = c(-300, 300, -300, 300),
                                     axis_breaks  = c(-300, -150, 0, 150, 300),
                                     title = NULL,
                                     show_legend = TRUE,
                                     outline_col = "#B64249") {

  group_name <- match.arg(group_name)

  h <- hulls_df %>% dplyr::filter(Group == group_name)
  if (!nrow(h)) return(ggplot2::ggplot() + ggplot2::ggtitle(paste0(group_name, " — EMPTY")))

  samples <- sort(unique(h$SampleID))
  sfx <- sub("^.*S(\\d)$", "\\1", samples)
  style_by_suffix <- c("1" = "solid", "2" = "longdash", "3" = "1212")
  lt_vals <- style_by_suffix[sfx]
  lt_vals[is.na(lt_vals)] <- "solid"
  lt_map <- setNames(lt_vals, samples)

  fill_cols <- setNames(rep(outline_col, length(samples)), samples)

  xlim <- fixed_limits[1:2]
  ylim <- fixed_limits[3:4]

  p <- ggplot2::ggplot(h, ggplot2::aes(x = RelX, y = RelY, group = SampleID)) +
    ggplot2::geom_polygon(ggplot2::aes(fill = SampleID), alpha = 0.22, color = NA) +
    ggplot2::geom_path(ggplot2::aes(linetype = SampleID), color = outline_col, linewidth = 1.6) +
    ggplot2::scale_linetype_manual(values = lt_map, drop = FALSE) +
    ggplot2::scale_fill_manual(values = fill_cols, drop = FALSE) +
    ggplot2::scale_x_continuous(breaks = axis_breaks, expand = c(0, 0)) +
    ggplot2::scale_y_continuous(breaks = axis_breaks, expand = c(0, 0)) +
    ggplot2::labs(
      title = if (is.null(title)) paste0(group_name, " Sample Hull Overlays") else title,
      x = "Relative X Position (µm)",
      y = "Relative Y Position (µm)",
      linetype = "Sample",
      fill = "Sample"
    ) +
    ggplot2::theme_classic(base_size = 22) +
    ggplot2::theme(
      legend.position = if (show_legend) "right" else "none",
      axis.title = ggplot2::element_text(size = 20),
      axis.text  = ggplot2::element_text(size = 18),
      plot.title = ggplot2::element_text(size = 22, face = "bold"),
      plot.margin = ggplot2::margin(10, 10, 10, 10)
    )

  # one coord system, with equal scaling
  p + ggplot2::coord_fixed(ratio = 1, xlim = xlim, ylim = ylim, clip = "off")
}

sample_hulls_MDA <- build_sample_hulls(Combined_Positions_Relative, cell_filter = "MDAMB231")
sample_hulls_Myo <- build_sample_hulls(Combined_Positions_Relative, cell_filter = "Myoepithelial")

# MDAMB231 overlays (±300)
p_den1_hull_MDA <- plot_group_hull_overlays(
  sample_hulls_MDA, "Den1",
  fixed_limits = c(-300, 300, -300, 300),
  axis_breaks  = c(-300, -150, 0, 150, 300),
  title = "MDAMB231 — Low Den (Sample Hull Overlays)",
  show_legend = TRUE,
  outline_col = "#B64249"
)

p_den10_hull_MDA <- plot_group_hull_overlays(
  sample_hulls_MDA, "Den10",
  fixed_limits = c(-300, 300, -300, 300),
  axis_breaks  = c(-300, -150, 0, 150, 300),
  title = "MDAMB231 — High Den (Sample Hull Overlays)",
  show_legend = TRUE,
  outline_col = "#B64249"
)

ggsave("ConvexHulls_MDAMB231_Den1_Overlays.png",  p_den1_hull_MDA,  width = 7, height = 6, dpi = 600)

ggsave("ConvexHulls_MDAMB231_Den10_Overlays.png", p_den10_hull_MDA, width = 7, height = 6, dpi = 600)

# Myoep overlays (±600)
p_den1_hull_Myo <- plot_group_hull_overlays(
  sample_hulls_Myo, "Den1",
  fixed_limits = c(-600, 600, -600, 600),
  axis_breaks  = c(-600, -300, 0, 300, 600),
  title = "Myoepithelial — Low Den (Sample Hull Overlays)",
  show_legend = TRUE,
  outline_col = "#796089"
)

p_den10_hull_Myo <- plot_group_hull_overlays(
  sample_hulls_Myo, "Den10",
  fixed_limits = c(-600, 600, -600, 600),
  axis_breaks  = c(-600, -300, 0, 300, 600),
  title = "Myoepithelial — High Den (Sample Hull Overlays)",
  show_legend = TRUE,
  outline_col = "#796089"
)

ggsave("ConvexHulls_Myoepithelial_Den1_Overlays.png",  p_den1_hull_Myo,  width = 7, height = 6, dpi = 600)

ggsave("ConvexHulls_Myoepithelial_Den10_Overlays.png", p_den10_hull_Myo, width = 7, height = 6, dpi = 600)

p_den1_hull_MDA
p_den10_hull_MDA
p_den1_hull_Myo
p_den10_hull_Myo
```

##### Stats
```{r}
SampleHullAreas <- Combined_Positions_Relative %>%
  dplyr::filter(CellTypeID %in% c("MDAMB231", "Myoepithelial")) %>%
  dplyr::filter(Group %in% c("Den1", "Den10")) %>%
  dplyr::group_by(CellTypeID, SampleID, Group) %>%
  dplyr::summarise(
    hull_area = {
      pts <- cbind(RelX, RelY)
      pts <- pts[stats::complete.cases(pts), , drop = FALSE]
      if (nrow(pts) < 3) NA_real_ else geometry::convhulln(pts, "FA")$area
    },
    .groups = "drop"
  )

SampleHullAreas

# Raw Welch t-tests 
stats::t.test(hull_area ~ Group, data = SampleHullAreas %>% dplyr::filter(CellTypeID == "MDAMB231"))

stats::t.test(hull_area ~ Group, data = SampleHullAreas %>% dplyr::filter(CellTypeID == "Myoepithelial"))
```

```{r}
t.test(hull_area ~ Group, data = SampleHullAreas %>% filter(CellTypeID == "MDAMB231"))

t.test(hull_area ~ Group, data = SampleHullAreas %>% filter(CellTypeID == "Myoepithelial"))
```

##### Per sample quant
```{r}
star_from_p <- function(p) {
  if (is.na(p)) return(NA_character_)
  if (p < 0.0001) return("****")
  if (p < 0.001)  return("***")
  if (p < 0.01)   return("**")
  if (p < 0.05)   return("*")
  "ns"
}

plot_hull_area_box <- function(SampleHullAreas, cell_type,
                               out_png,
                               title = cell_type,
                               bracket_y = NULL) {

  df_plot <- SampleHullAreas %>%
    filter(CellTypeID == cell_type) %>%
    filter(!is.na(hull_area), is.finite(hull_area), hull_area > 0) %>%
    mutate(Group = factor(Group, levels = c("Den1", "Den10")))

  # guardrails so t.test doesn't error
  if (n_distinct(df_plot$Group) < 2 || any(table(df_plot$Group) < 2)) {
    stop("Not enough samples per group for ", cell_type, " (need both groups and >=2 per group).")
  }

  tt  <- t.test(hull_area ~ Group, data = df_plot, var.equal = FALSE)
  sig <- star_from_p(tt$p.value)

  data_max <- max(df_plot$hull_area, na.rm = TRUE)
  base_y   <- if (!is.null(bracket_y)) bracket_y else data_max * 1.06

  stat_df_hull <- tibble(
    group1     = "Den1",
    group2     = "Den10",
    y.position = base_y,
    label      = sig
  ) %>%
    filter(!is.na(label), label != "ns")

  ylim_top_hull <- if (nrow(stat_df_hull) > 0) {
    max(stat_df_hull$y.position, na.rm = TRUE) * 1.08
  } else {
    data_max * 1.05
  }

  step       <- if (ylim_top_hull <= 2000) 250 else 500
  ymax_break <- ceiling(ylim_top_hull / step) * step
  y_breaks   <- seq(0, ymax_break, by = step)

  p_box <- ggplot(df_plot, aes(x = Group, y = hull_area)) +
    geom_boxplot(
      aes(fill = Group),
      width = 0.55,
      outlier.shape = NA,
      alpha = 0.85,
      linewidth = 0.8
    ) +
    geom_jitter(
      aes(color = Group),
      width = 0.3,
      size = 3,
      alpha = 0.8,
      show.legend = FALSE
    ) +
    scale_x_discrete(labels = c("Den1" = "Low Den", "Den10" = "High Den")) +
    scale_fill_manual(values = group_colors, drop = FALSE) +
    scale_color_manual(values = point_colors, drop = FALSE) +
    scale_y_continuous(breaks = y_breaks, expand = c(0, 0)) +
    coord_cartesian(ylim = c(0, ylim_top_hull), clip = "off") +
    {
      if (nrow(stat_df_hull) > 0) {
        stat_pvalue_manual(
          stat_df_hull,
          xmin = "group1",
          xmax = "group2",
          label = "label",
          y.position = "y.position",
          tip.length = 0.03,
          size = 14,
          bracket.size = 0.9
        )
      } else NULL
    } +
    labs(
      title = title,
      x = "Cancer Density",
      y = expression("Migration Territory ("*mu*m^2*")")
    ) +
    theme_classic(base_size = 22) +
    theme(
      axis.line    = element_line(linewidth = 0.8, color = "black"),
      axis.text.y  = element_text(size = 18),
      axis.title.y = element_text(size = 20),
      axis.text.x  = element_text(size = 18, hjust = 0.5),
      axis.title.x = element_text(size = 20),
      plot.title   = element_text(size = 18, hjust = 0.5),
      legend.position = "none",
      plot.margin = margin(10, 10, 10, 10)
    )

  ggsave(out_png, p_box, width = 4.5, height = 6, dpi = 600)
  p_box
}

# Make both plots
p_box_MDA <- plot_hull_area_box(
  SampleHullAreas, "MDAMB231",
  out_png = "HullArea_MDAMB231_D1vD10.png",
  title   = "MDAMB231"
)

p_box_Myo <- plot_hull_area_box(
  SampleHullAreas, "Myoepithelial",
  out_png = "HullArea_Myoepithelial_D1vD10.png",
  title   = "Myoepithelial"
)

p_box_MDA
p_box_Myo
```

### 3.) Total Distance Traveled
```{r}
CP_num_len <- Combined_Positions %>%
  mutate(
    TimeNum = as.integer(Time),
    X       = as.numeric(X_Position),
    Y       = as.numeric(Y_Position)
  ) %>%
  arrange(SampleID, UniqueID, TimeNum)

# Track length per track (UniqueID)
TrackLengths <- CP_num_len %>%
  group_by(SampleID, UniqueID, Group, CellTypeID) %>%
  summarise(
    TrackLength = {
      x <- X
      y <- Y
      ok <- complete.cases(x, y)
      x <- x[ok]; y <- y[ok]
      if (length(x) < 2) {
        NA_real_
      } else {
        sum(sqrt(diff(x)^2 + diff(y)^2))
      }
    },
    n_points = sum(complete.cases(X, Y)),
    .groups = "drop"
  )

df_len <- TrackLengths %>%
  filter(
    CellTypeID %in% c("MDAMB231", "Myoepithelial"),
    Group %in% c("Den1", "Den10"),
    !is.na(TrackLength),
    is.finite(TrackLength)
  ) %>%
  mutate(
    Group      = factor(Group, levels = c("Den1", "Den10")),
    CellTypeID = factor(CellTypeID, levels = c("MDAMB231", "Myoepithelial"))
  )

# Welch t-tests (per cell type)
star_from_p <- function(p) {
  if (is.na(p)) return(NA_character_)
  if (p < 0.0001) return("****")
  if (p < 0.001)  return("***")
  if (p < 0.01)   return("**")
  if (p < 0.05)   return("*")
  "ns"
}

welch_tracklength_by_celltype <- function(df) {

  res <- df %>%
    group_by(CellTypeID) %>%
    group_modify(function(d, key) {

      d <- d %>%
        tidyr::drop_na(TrackLength, Group) %>%
        mutate(Group = factor(Group, levels = c("Den1", "Den10")))

      # require both groups + >=2 tracks per group
      if (n_distinct(d$Group) < 2 || any(table(d$Group) < 2)) {
        return(tibble(
          comparison = "Den1-Den10",
          direction  = NA_character_,
          p_value    = NA_real_,
          sig        = NA_character_,
          statement  = paste0(as.character(key$CellTypeID), ": Not enough tracks in one group")
        ))
      }

      tt <- t.test(TrackLength ~ Group, data = d, var.equal = FALSE)
      p  <- tt$p.value

      means <- tapply(d$TrackLength, d$Group, mean, na.rm = TRUE)
      dir_txt <- if (means["Den1"] > means["Den10"]) "Den1 > Den10" else "Den10 > Den1"

      tibble(
        comparison = "Den1-Den10",
        direction  = dir_txt,
        p_value    = p,
        sig        = star_from_p(p),
        statement  = paste0(
          as.character(key$CellTypeID),
          " TrackLength: ", dir_txt,
          " (Welch t-test, p = ", signif(p, 3), ", ", star_from_p(p), ")"
        )
      )
    }) %>%
    ungroup()

  res
}

tracklen_pairwise <- welch_tracklength_by_celltype(df_len)
```

#### Stats
```{r}
tracklen_pairwise
tracklen_pairwise$statement
```

#### Viz
```{r}
fill_cols <- group_colors
point_cols <- fill_cols  #do this so we can set the alpha different

ymax <- max(df_len$TrackLength, na.rm = TRUE)
ymax_ceil <- ceiling(ymax / 250) * 250
step <- if (ymax_ceil <= 2000) 250 else 500
y_breaks <- seq(0, ymax_ceil, by = step)

sig_rows <- tracklen_pairwise %>%
  filter(!is.na(sig), sig != "ns") %>%
  mutate(
    group1 = "Den1",
    group2 = "Den10"
  )

# bracket heights per facet
base_y <- ymax_ceil * 0.80
step_y <- ymax_ceil * 0.08

stat_df_len <- sig_rows %>%
  group_by(CellTypeID) %>%
  mutate(
    y.position = base_y + step_y * (row_number() - 1),
    label      = sig
  ) %>%
  ungroup() %>%
  select(CellTypeID, group1, group2, y.position, label)

ylim_top_len <- if (nrow(stat_df_len) > 0) {
  max(stat_df_len$y.position, na.rm = TRUE) * 1.10
} else {
  ymax_ceil * 1.10
}

p_len <- ggplot(df_len, aes(x = Group, y = TrackLength)) +
  geom_violin(
    aes(fill = Group),
    trim  = FALSE,
    alpha = 0.35,
    color = NA,
    width = 0.95
  ) +
  geom_boxplot(
    aes(fill = Group),
    width         = 0.22,
    outlier.shape = NA,
    alpha         = 0.85,
    linewidth     = 0.8
  ) +
  geom_jitter(
    aes(color = Group),
    width = 0.15,
    size  = 2,
    alpha = 0.55,
    show.legend = FALSE
  ) +
  facet_grid(. ~ CellTypeID) +
  scale_fill_manual(values = fill_cols, drop = FALSE) +
  scale_color_manual(values = point_cols, drop = FALSE) +
  scale_y_continuous(breaks = y_breaks, expand = c(0, 0)) +
  scale_x_discrete(labels = c("Den1" = "Low Den", "Den10" = "High Den")) +
  coord_cartesian(ylim = c(0, ylim_top_len)) +
  {
    if (nrow(stat_df_len) > 0) {
      stat_pvalue_manual(
        stat_df_len,
        xmin = "group1",
        xmax = "group2",
        label = "label",
        y.position = "y.position",
        tip.length   = 0.03,
        size         = 6,
        bracket.size = 0.9
      )
    } else {
      NULL
    }
  } +
  labs(
    title = NULL,
    x     = "Cancer Density",
    y     = expression("Total Distance Traveled ("*mu*"m)")
  ) +
  theme_classic(base_size = 20) +
  theme(
    axis.line    = element_line(linewidth = 0.9, color = "black"),
    axis.text.y  = element_text(size = 18),
    axis.title.y = element_text(size = 20),
    axis.text.x  = element_text(size = 16, angle = 0, hjust = 0.5),
    axis.title.x = element_text(size = 16),
    legend.position = "none",
    strip.background = element_blank(),
    strip.text.x = element_text(size = 20),
    panel.spacing.x = unit(2, "lines")
  )

ggsave(
  filename = "TrackLength_ViolinBox_StarsOnly_Den1_Den10.png",
  plot     = p_len,
  width    = 6.5,
  height   = 4.7,
  dpi      = 600
)

p_len
```


### 4.) Normalized Distance (Distance ÷ Frames) Accounts for tracks truncated from cell divisions and cells that go out of frame.
```{r}
CP_num_df <- Combined_Positions %>%
  mutate(
    TimeNum = suppressWarnings(as.integer(Time)),
    X       = suppressWarnings(as.numeric(X_Position)),
    Y       = suppressWarnings(as.numeric(Y_Position))
  ) %>%
  arrange(SampleID, UniqueID, TimeNum)

DistPerFrame_df <- CP_num_df %>%
  filter(CellTypeID %in% c("MDAMB231", "Myoepithelial")) %>%
  group_by(SampleID, UniqueID, Group, CellTypeID) %>%
  summarise(
    n_frames_visible = n_distinct(TimeNum[complete.cases(X, Y)]),

    TrackLength = {
      d <- data.frame(TimeNum = TimeNum, X = X, Y = Y) %>%
        filter(complete.cases(TimeNum, X, Y)) %>%
        arrange(TimeNum)

      if (nrow(d) < 2) NA_real_ else sum(sqrt(diff(d$X)^2 + diff(d$Y)^2))
    },
    .groups = "drop"
  ) %>%
  filter(!is.na(TrackLength), n_frames_visible >= 2) %>%
  mutate(
    Group = factor(Group, levels = c("Den1", "Den10")),
    CellTypeID = factor(CellTypeID, levels = c("MDAMB231", "Myoepithelial")),
    Dist_um_per_frame = TrackLength / (n_frames_visible - 1)
  )

df_dist_frame <- DistPerFrame_df
```

#### Stats
```{r}
# Welch t-test per cell type
star_from_p <- function(p) {
  if (is.na(p)) return(NA_character_)
  if (p < 0.0001) return("****")
  if (p < 0.001)  return("***")
  if (p < 0.01)   return("**")
  if (p < 0.05)   return("*")
  "ns"
}

pairwise_dist_frame_den <- function(df) {

  df %>%
    group_by(CellTypeID) %>%
    group_modify(function(d, key) {

      d <- d %>%
        tidyr::drop_na(Dist_um_per_frame, Group) %>%
        mutate(Group = factor(Group, levels = c("Den1", "Den10")))

      # require both groups + >=2 tracks per group
      if (n_distinct(d$Group) < 2 || any(table(d$Group) < 2)) {
        return(tibble(
          comparison = "Den1-Den10",
          direction  = NA_character_,
          p_value    = NA_real_,
          sig        = NA_character_,
          statement  = paste0(as.character(key$CellTypeID), ": Not enough tracks in one group")
        ))
      }

      tt <- t.test(Dist_um_per_frame ~ Group, data = d, var.equal = FALSE)
      p  <- tt$p.value

      means <- tapply(d$Dist_um_per_frame, d$Group, mean, na.rm = TRUE)
      dir_txt <- if (means["Den1"] > means["Den10"]) "Den1 > Den10" else "Den10 > Den1"

      tibble(
        comparison = "Den1-Den10",
        direction  = dir_txt,
        p_value    = p,
        sig        = star_from_p(p),
        statement  = paste0(
          as.character(key$CellTypeID),
          " Distance/frame: ", dir_txt,
          " (Welch t-test, p = ", signif(p, 3), ", ", star_from_p(p), ")"
        )
      )
    }) %>%
    ungroup()
}

dist_frame_pairwise <- pairwise_dist_frame_den(df_dist_frame)

dist_frame_pairwise
dist_frame_pairwise$statement
```

#### Viz
```{r}
fill_cols  <- group_colors
point_cols <- point_colors

sig_rows <- dist_frame_pairwise %>%
  filter(!is.na(sig), sig != "ns") %>%
  mutate(group1 = "Den1", group2 = "Den10")

if (nrow(df_dist_frame) == 0) {
  stop("df_dist_frame is empty. Check Group assignment + CellTypeID filtering upstream.")
}

data_max <- max(df_dist_frame$Dist_um_per_frame, na.rm = TRUE)
if (!is.finite(data_max)) {
  stop("Dist_um_per_frame is all NA or non-finite. Check numeric conversion of X/Y/Time.")
}

# bracket heights per facet
base_y <- data_max * 1.2
step_y <- data_max * 0.08

stat_df_dist <- sig_rows %>%
  group_by(CellTypeID) %>%
  mutate(
    y.position = base_y + step_y * (row_number() - 1),
    label      = sig
  ) %>%
  ungroup() %>%
  select(CellTypeID, group1, group2, y.position, label)

ylim_top_dist <- if (nrow(stat_df_dist) > 0) {
  max(stat_df_dist$y.position, na.rm = TRUE) * 1.10
} else {
  data_max * 1.20
}

# y breaks
step <- if (ylim_top_dist <= 10) 1 else if (ylim_top_dist <= 25) 2.5 else 5
ymax_break <- ceiling(ylim_top_dist / step) * step
y_breaks <- seq(0, ymax_break, by = step)

p_dist_frame <- ggplot(df_dist_frame, aes(x = Group, y = Dist_um_per_frame)) +
  geom_violin(aes(fill = Group), trim = FALSE, alpha = 0.35, color = NA, width = 0.95) +
  geom_boxplot(aes(fill = Group), width = 0.22, outlier.shape = NA, alpha = 0.85, linewidth = 0.8) +
  geom_jitter(aes(color = Group), width = 0.15, size = 2, alpha = 0.55, show.legend = FALSE) +
  facet_grid(. ~ CellTypeID) +
  scale_fill_manual(values = fill_cols, drop = FALSE) +
  scale_color_manual(values = point_cols, drop = FALSE) +
  scale_y_continuous(breaks = y_breaks, expand = c(0, 0)) +
  scale_x_discrete(labels = c("Den1" = "Low Den", "Den10" = "High Den")) +
  coord_cartesian(ylim = c(0, ylim_top_dist)) +
  {
    if (nrow(stat_df_dist) > 0) {
      stat_pvalue_manual(
        stat_df_dist,
        xmin = "group1",
        xmax = "group2",
        label = "label",
        y.position = "y.position",
        tip.length = 0.03,
        size = 14,
        bracket.size = 0.9
      )
    } else NULL
  } +
  labs(
    title = NULL,
    x     = "Cancer Density",
    y     = "Normalized Distance \nTraveled (µm / Frame)"
  ) +
  theme_classic(base_size = 20) +
  theme(
    axis.line    = element_line(linewidth = 0.9, color = "black"),
    axis.text.y  = element_text(size = 18),
    axis.title.y = element_text(size = 20, angle = 90, vjust = 0.45),
    axis.text.x  = element_text(size = 16, hjust = 0.5),
    axis.title.x = element_text(size = 16),
    legend.position = "none",
    strip.background = element_blank(),
    strip.text.x = element_text(size = 20),
    panel.spacing.x = unit(2, "lines"),
    plot.margin = margin(t = 10, r = 10, b = 10, l = 28, unit = "pt")
  )

ggsave(
  filename = "DistancePerVisibleFrame_ViolinBox_StarsOnly_Den1_Den10.png",
  plot     = p_dist_frame,
  width    = 6.5,
  height   = 5,
  dpi      = 600
)

p_dist_frame
```

### 5.) Net Displacement
```{r}
fill_cols  <- group_colors
point_cols <- point_colors

# Build per-track net displacement
disp_df <- Combined_Positions %>%
  mutate(
    TimeNum = as.integer(Time),
    X       = as.numeric(X_Position),
    Y       = as.numeric(Y_Position)
  ) %>%
  filter(
    is.finite(TimeNum), is.finite(X), is.finite(Y),
    Group %in% c("Den1", "Den10"),
    CellTypeID %in% c("MDAMB231", "Myoepithelial"),
    CellTypeID != "Missing_ID"
  ) %>%
  arrange(SampleID, UniqueID, TimeNum) %>%
  group_by(SampleID, UniqueID) %>%
  summarise(
    Group      = first(Group),
    CellTypeID = first(CellTypeID),

    # sanity checks (drop broken tracks)
    group_n    = n_distinct(Group),
    celltype_n = n_distinct(CellTypeID),

    x0 = first(X),
    y0 = first(Y),
    x1 = last(X),
    y1 = last(Y),

    Displacement_um = sqrt((x1 - x0)^2 + (y1 - y0)^2),
    n_points = n(),
    .groups = "drop"
  ) %>%
  filter(group_n == 1, celltype_n == 1, n_points >= 2) %>%
  select(-group_n, -celltype_n) %>%
  mutate(
    Group      = factor(Group, levels = c("Den1", "Den10")),
    CellTypeID = factor(CellTypeID, levels = c("MDAMB231", "Myoepithelial"))
  )

# optional count check
disp_df %>% count(CellTypeID, Group)
```

#### Stats
```{r}
# Welch t-test per cell type
star_from_p <- function(p) {
  if (is.na(p)) return(NA_character_)
  if (p < 0.0001) return("****")
  if (p < 0.001)  return("***")
  if (p < 0.01)   return("**")
  if (p < 0.05)   return("*")
  "ns"
}

welch_displacement_by_celltype <- function(df) {

  df %>%
    group_by(CellTypeID) %>%
    group_modify(function(d, key) {

      d <- d %>%
        tidyr::drop_na(Displacement_um, Group) %>%
        mutate(Group = factor(Group, levels = c("Den1", "Den10")))

      # require both groups + >=2 tracks per group
      if (n_distinct(d$Group) < 2 || any(table(d$Group) < 2)) {
        return(tibble(
          comparison = "Den1-Den10",
          direction  = NA_character_,
          p_value    = NA_real_,
          sig        = NA_character_,
          statement  = paste0(as.character(key$CellTypeID), ": Not enough tracks in one group")
        ))
      }

      tt <- t.test(Displacement_um ~ Group, data = d, var.equal = FALSE)
      p  <- tt$p.value

      means <- tapply(d$Displacement_um, d$Group, mean, na.rm = TRUE)
      dir_txt <- if (means["Den1"] > means["Den10"]) "Den1 > Den10" else "Den10 > Den1"

      tibble(
        comparison = "Den1-Den10",
        direction  = dir_txt,
        p_value    = p,
        sig        = star_from_p(p),
        statement  = paste0(
          as.character(key$CellTypeID),
          " Net Displacement: ", dir_txt,
          " (Welch t-test, p = ", signif(p, 3), ", ", star_from_p(p), ")"
        )
      )
    }) %>%
    ungroup()
}

disp_pairwise <- welch_displacement_by_celltype(disp_df)

disp_pairwise
disp_pairwise$statement
```

#### Viz
```{r}
# Viz (violin + box + jitter; stars only; faceted by cell type)

sig_rows <- disp_pairwise %>%
  filter(!is.na(sig), sig != "ns") %>%
  mutate(group1 = "Den1", group2 = "Den10")

if (nrow(disp_df) == 0) {
  stop("disp_df is empty. Check Group + CellTypeID filtering upstream.")
}

data_max <- max(disp_df$Displacement_um, na.rm = TRUE)
if (!is.finite(data_max)) {
  stop("Displacement_um is all NA or non-finite.")
}

# "nice" step sizes for displacement
step <- if (data_max <= 500) 50 else if (data_max <= 1000) 100 else if (data_max <= 2000) 250 else 500
ymax_ceil <- ceiling(data_max / step) * step

# bracket heights per facet
base_y <- ymax_ceil * 0.80
step_y <- ymax_ceil * 0.08

stat_df_disp <- sig_rows %>%
  group_by(CellTypeID) %>%
  mutate(
    y.position = base_y + step_y * (row_number() - 1),
    label      = sig
  ) %>%
  ungroup() %>%
  select(CellTypeID, group1, group2, y.position, label)

ylim_top_disp <- if (nrow(stat_df_disp) > 0) {
  max(stat_df_disp$y.position, na.rm = TRUE) * 1.10
} else {
  ymax_ceil * 1.10
}

# extend breaks to top
ymax_break <- ceiling(ylim_top_disp / step) * step
y_breaks   <- seq(0, ymax_break, by = step)

p_disp <- ggplot(disp_df, aes(x = Group, y = Displacement_um)) +
  geom_violin(aes(fill = Group), trim = FALSE, alpha = 0.35, color = NA, width = 0.95) +
  geom_boxplot(aes(fill = Group), width = 0.22, outlier.shape = NA, alpha = 0.85, linewidth = 0.8) +
  geom_jitter(aes(color = Group), width = 0.15, size = 2, alpha = 0.55, show.legend = FALSE) +
  facet_grid(. ~ CellTypeID) +
  scale_fill_manual(values = fill_cols, drop = FALSE) +
  scale_color_manual(values = point_cols, drop = FALSE) +
  scale_y_continuous(breaks = y_breaks, expand = c(0, 0)) +
  scale_x_discrete(labels = c("Den1" = "Low Den", "Den10" = "High Den")) +
  coord_cartesian(ylim = c(0, ylim_top_disp)) +
  {
    if (nrow(stat_df_disp) > 0) {
      ggpubr::stat_pvalue_manual(
        stat_df_disp,
        xmin = "group1",
        xmax = "group2",
        label = "label",
        y.position = "y.position",
        tip.length   = 0.03,
        size         = 6,
        bracket.size = 0.9
      )
    } else NULL
  } +
  labs(
    title = NULL,
    x = "Cancer Density",
    y = expression("Net Displacement ("*mu*m*")")
  ) +
  theme_classic(base_size = 20) +
  theme(
    axis.line    = element_line(linewidth = 0.9, color = "black"),
    axis.text.y  = element_text(size = 18),
    axis.title.y = element_text(size = 20),
    axis.text.x  = element_text(size = 16, hjust = 0.5),
    axis.title.x = element_text(size = 16),
    legend.position = "none",
    strip.background = element_blank(),
    strip.text.x = element_text(size = 20),
    panel.spacing.x = unit(2, "lines")
  )

ggsave(
  filename = "NetDisplacement_ViolinBox_StarsOnly_Den1_Den10.png",
  plot     = p_disp,
  width    = 6.5,
  height   = 4.7,
  dpi      = 600
)

p_disp
```


### 6.) MSD
```{r}
frame_interval_min <- 10

prep_tracks <- function(df) {
  df %>%
    mutate(
      Time       = suppressWarnings(readr::parse_number(as.character(Time))),
      X_Position = suppressWarnings(readr::parse_number(as.character(X_Position))),
      Y_Position = suppressWarnings(readr::parse_number(as.character(Y_Position)))
    ) %>%
    filter(is.finite(Time), is.finite(X_Position), is.finite(Y_Position)) %>%
    arrange(UniqueID, Time) %>%
    distinct(UniqueID, Time, .keep_all = TRUE)
}

tasd_per_track <- function(track_df) {
  x <- track_df$X_Position
  y <- track_df$Y_Position
  n <- length(x)
  if (n < 2) return(tibble::tibble(tau_frames = integer(), msd_track = numeric()))

  k_vec <- seq_len(n - 1)
  msd_k <- vapply(k_vec, function(k) {
    dx <- x[(1 + k):n] - x[1:(n - k)]
    dy <- y[(1 + k):n] - y[1:(n - k)]
    mean(dx*dx + dy*dy, na.rm = TRUE)
  }, numeric(1))

  tibble::tibble(tau_frames = k_vec, msd_track = msd_k)
}

calc_tasd_unbalanced <- function(df, frame_interval_min = 10, min_tracks_for_sem = 3) {
  df <- df %>%
    prep_tracks() %>%
    group_by(UniqueID) %>%
    arrange(Time, .by_group = TRUE)

  if (nrow(df) == 0) {
    return(tibble::tibble(
      tau_frames = integer(),
      tau_min    = numeric(),
      tasd       = numeric(),
      sem        = numeric(),
      n_tracks   = integer()
    ))
  }

  per_track <- df %>%
    group_split(UniqueID, .keep = FALSE) %>%
    purrr::map_dfr(tasd_per_track, .id = "track_index")

  per_track %>%
    group_by(tau_frames) %>%
    summarise(
      tasd     = mean(msd_track, na.rm = TRUE),
      sem      = sd(msd_track, na.rm = TRUE) / sqrt(sum(is.finite(msd_track))),
      n_tracks = sum(is.finite(msd_track)),
      .groups  = "drop"
    ) %>%
    mutate(
      tau_min = tau_frames * frame_interval_min,
      tau_hr  = tau_min / 60,                          # <<< NEW: hours
      sem     = ifelse(n_tracks >= min_tracks_for_sem, sem, NA_real_)
    ) %>%
    arrange(tau_frames)
}

compute_by_celltype <- function(dat,
                                celltype,
                                groups = c("Den1","Den10"),
                                frame_interval_min = 10,
                                min_tracks_for_sem = 3) {
  purrr::map_dfr(groups, function(g) {
    subset_df <- dat %>%
      filter(CellTypeID == celltype, Group == g) %>%
      select(UniqueID, Time, X_Position, Y_Position)

    out <- calc_tasd_unbalanced(
      subset_df,
      frame_interval_min = frame_interval_min,
      min_tracks_for_sem = min_tracks_for_sem
    )
    out$Group      <- g
    out$CellTypeID <- celltype
    out
  })
}

trim_to_common_tau <- function(summary_df, groups = c("Den1","Den10")) {
  df <- summary_df %>% filter(Group %in% groups)

  max_tau_by_group <- df %>%
    group_by(CellTypeID, Group) %>%
    summarise(
      max_tau = max(tau_frames[n_tracks >= 1], na.rm = TRUE),
      .groups = "drop"
    )

  common_tau_by_ct <- max_tau_by_group %>%
    group_by(CellTypeID) %>%
    summarise(common_tau = min(max_tau, na.rm = TRUE), .groups = "drop")

  df %>%
    left_join(common_tau_by_ct, by = "CellTypeID") %>%
    filter(is.finite(common_tau), tau_frames <= common_tau) %>%
    select(-common_tau)
}

# Compute summaries
mda_sum <- compute_by_celltype(
  Combined_Positions,
  celltype = "MDAMB231",
  groups = c("Den1","Den10"),
  frame_interval_min = frame_interval_min,
  min_tracks_for_sem = 3
)

myo_sum <- compute_by_celltype(
  Combined_Positions,
  celltype = "Myoepithelial",
  groups = c("Den1","Den10"),
  frame_interval_min = frame_interval_min,
  min_tracks_for_sem = 3
)

all_sum <- bind_rows(mda_sum, myo_sum) %>%
  mutate(
    Group      = factor(Group, levels = c("Den1","Den10")),
    CellTypeID = factor(CellTypeID, levels = c("MDAMB231","Myoepithelial"))
  )

# Trim to common tau per cell type so Den1 and Den10 have identical x-range within each facet
all_sum_trim <- trim_to_common_tau(all_sum, groups = c("Den1","Den10"))

# y-limit based on observed (tasd + sem)
y_max <- max(all_sum_trim$tasd + ifelse(is.finite(all_sum_trim$sem), all_sum_trim$sem, 0), na.rm = TRUE)
y_max <- ceiling(y_max / 1000) * 1000

# Ribbon only where SEM exists
all_ribbon <- all_sum_trim %>% filter(is.finite(sem))

xmax_hr <- max(all_sum_trim$tau_hr, na.rm = TRUE)
x_breaks_hr <- seq(0, ceiling(xmax_hr / 6) * 6, by = 6)

all_sum <- bind_rows(mda_sum, myo_sum) %>%
  mutate(
    Group = factor(
      Group,
      levels = c("Den1", "Den10"),
      labels = c("Low Den", "High Den")
    ),
    CellTypeID = factor(CellTypeID, levels = c("MDAMB231","Myoepithelial"))
  )

p_tasd <- ggplot(all_sum_trim, aes(x = tau_hr, y = tasd, color = Group, fill = Group)) +
  geom_ribbon(
    data = all_ribbon,
    aes(ymin = pmax(tasd - sem, 0), ymax = tasd + sem),
    alpha = 0.18,
    color = NA
  ) +
  geom_line(linewidth = 0.9, na.rm = TRUE) +
  geom_point(size = 0.5, alpha = 0.6, na.rm = TRUE) +
  facet_grid(. ~ CellTypeID) +
  scale_color_manual(
  values = group_colors,
  breaks = c("Den1", "Den10"),
  labels = c("Low Den", "High Den"),
  name   = "Cancer \nDensity",
  drop   = FALSE
) +
scale_fill_manual(
  values = group_colors,
  breaks = c("Den1", "Den10"),
  labels = c("Low Den", "High Den"),
  name   = "Cancer \nDensity",
  drop   = FALSE
) +
  scale_x_continuous(
    limits = c(0, NA),
    breaks = x_breaks_hr,
    expand = c(0, 0)
  ) +
  scale_y_continuous(limits = c(0, y_max), expand = expansion(mult = c(0, 0.02))) +
  labs(
    title = NULL,
    x = "Lag τ (hours)",
    y = expression("MSD ("*µ*m^2*")")
  ) +
  theme_classic(base_size = 22) +
  theme(
    plot.title       = element_text(size = 24),
    legend.position  = "right",
    legend.title = element_text(size = 18, margin = margin(b = 4)),
    legend.text = element_text(size = 16),
    legend.key.height = unit(10, "pt"),
    legend.spacing.y  = unit(2, "pt"),
    strip.background = element_blank(),
    strip.text.x     = element_text(size = 20),
    axis.line        = element_line(linewidth = 1.1, color = "black"),
    axis.text        = element_text(size = 18),
    axis.title       = element_text(size = 20),
    panel.spacing.x  = grid::unit(2, "cm")
  )


ggsave("TASD_Den1_Den10_MDAMB231_Myo_sharedY.png", 
       p_tasd, 
       width = 12.5, 
       height = 6, 
       dpi = 600)

p_tasd
```

#### Alpha calculation
```{r}
fit_range_hr <- c(0.83, 8.33)   # 50 to 500 minutes expressed in hours

get_alpha <- function(df,
                      tau_col = "tau_hr",         
                      msd_col = "tasd",
                      tau_range = c(0.83, 8.33),
                      min_pts = 3) {

  stopifnot(tau_col %in% names(df), msd_col %in% names(df))

  df_filt <- df %>%
    dplyr::filter(
      is.finite(.data[[tau_col]]),
      .data[[tau_col]] >= tau_range[1],
      .data[[tau_col]] <= tau_range[2],
      is.finite(.data[[msd_col]]),
      .data[[msd_col]] > 0
    ) %>%
    dplyr::mutate(
      log_tau = log(.data[[tau_col]]),
      log_msd = log(.data[[msd_col]])
    )

  if (nrow(df_filt) == 0) {
    return(tibble::tibble(
      CellTypeID = character(),
      Group      = character(),
      n_pts      = integer(),
      alpha      = numeric(),
      r2         = numeric()
    ))
  }

  df_filt %>%
    dplyr::group_by(CellTypeID, Group) %>%
    tidyr::nest() %>%
    dplyr::mutate(
      n_pts  = purrr::map_int(data, nrow),
      model = purrr::map(data, ~ if (nrow(.x) >= min_pts) stats::lm(log_msd ~ log_tau, data = .x) else NULL),
      alpha = purrr::map_dbl(model, ~ if (is.null(.x)) NA_real_ else unname(stats::coef(.x)[2])),
      r2    = purrr::map_dbl(model, ~ if (is.null(.x)) NA_real_ else summary(.x)$r.squared)
    ) %>%
    dplyr::ungroup() %>%
    dplyr::select(CellTypeID, Group, n_pts, alpha, r2)
}

alpha_results <- get_alpha(
  all_sum_trim,
  tau_col   = "tau_hr",
  msd_col   = "tasd",
  tau_range = fit_range_hr,
  min_pts   = 3
)

alpha_results %>%
  dplyr::mutate(alpha = sprintf("%.2f", alpha))
```

## All not sig

### 7.) Travel Consistency
```{r}
if (!exists("TrackLengths")) {
  CP_num_len <- Combined_Positions %>%
    mutate(
      TimeNum = as.integer(Time),
      X       = as.numeric(X_Position),
      Y       = as.numeric(Y_Position)
    ) %>%
    arrange(SampleID, UniqueID, TimeNum)
  
  TrackLengths <- CP_num_len %>%
    group_by(SampleID, UniqueID, Group, CellTypeID) %>%
    summarise(
      TrackLength = {
        x <- X; y <- Y
        ok <- complete.cases(x, y)
        x <- x[ok]; y <- y[ok]
        if (length(x) < 2) NA_real_ else sum(sqrt(diff(x)^2 + diff(y)^2))
      },
      n_points = sum(complete.cases(X, Y)),
      .groups = "drop"
    )
}

if (!exists("disp_df")) {
  disp_df <- Combined_Positions %>%
    mutate(
      TimeNum = as.integer(Time),
      X       = as.numeric(X_Position),
      Y       = as.numeric(Y_Position)
    ) %>%
    filter(
      is.finite(TimeNum), is.finite(X), is.finite(Y),
      Group %in% c("Den1", "Den10"),
      CellTypeID %in% c("MDAMB231", "Myoepithelial"),
      CellTypeID != "Missing_ID"
    ) %>%
    arrange(SampleID, UniqueID, TimeNum) %>%
    group_by(SampleID, UniqueID) %>%
    summarise(
      Group      = first(Group),
      CellTypeID = first(CellTypeID),
      group_n    = n_distinct(Group),
      celltype_n = n_distinct(CellTypeID),
      x0 = first(X), y0 = first(Y),
      x1 = last(X),  y1 = last(Y),
      Displacement_um = sqrt((x1 - x0)^2 + (y1 - y0)^2),
      n_points = n(),
      .groups = "drop"
    ) %>%
    filter(group_n == 1, celltype_n == 1, n_points >= 2) %>%
    select(-group_n, -celltype_n) %>%
    mutate(
      Group      = factor(Group, levels = c("Den1", "Den10")),
      CellTypeID = factor(CellTypeID, levels = c("MDAMB231", "Myoepithelial"))
    )
}

# Merge + compute CI
CI_df <- TrackLengths %>%
  inner_join(
    disp_df %>% select(SampleID, UniqueID, Group, CellTypeID, Displacement_um),
    by = c("SampleID", "UniqueID", "Group", "CellTypeID")
  ) %>%
  mutate(
    TrackLength     = as.numeric(TrackLength),
    Displacement_um = as.numeric(Displacement_um),
    CI = if_else(
      is.finite(TrackLength) & TrackLength > 0 & is.finite(Displacement_um) & Displacement_um >= 0,
      Displacement_um / TrackLength,
      NA_real_
    )
  ) %>%
  filter(
    Group %in% c("Den1", "Den10"),
    CellTypeID %in% c("MDAMB231", "Myoepithelial"),
    is.finite(CI)
  ) %>%
  mutate(
    Group      = factor(Group, levels = c("Den1", "Den10")),
    CellTypeID = factor(CellTypeID, levels = c("MDAMB231", "Myoepithelial"))
  )
```

#### Stats
```{r}
pairwise_ci_den <- function(df) {

  star_from_p <- function(p) {
    if (is.na(p)) return(NA_character_)
    if (p <= 0.0001) return("****")
    if (p <= 0.001)  return("***")
    if (p <= 0.01)   return("**")
    if (p <= 0.05)   return("*")
    "ns"
  }

  df %>%
    group_by(CellTypeID) %>%
    group_modify(function(d, key) {

      d <- d %>%
        tidyr::drop_na(CI, Group) %>%
        mutate(Group = factor(Group, levels = c("Den1", "Den10")))

      # require both groups + >=2 tracks per group
      if (n_distinct(d$Group) < 2 || any(table(d$Group) < 2)) {
        return(tibble(
          comparison = "Den1-Den10",
          direction  = NA_character_,
          p_value    = NA_real_,
          sig        = NA_character_,
          statement  = paste0(as.character(key$CellTypeID), ": Not enough tracks in one group")
        ))
      }

      tt <- t.test(CI ~ Group, data = d, var.equal = FALSE)
      p  <- tt$p.value

      means <- tapply(d$CI, d$Group, mean, na.rm = TRUE)
      dir_txt <- if (means["Den1"] > means["Den10"]) "Den1 > Den10" else "Den10 > Den1"

      tibble(
        comparison = "Den1-Den10",
        direction  = dir_txt,
        p_value    = p,
        sig        = star_from_p(p),
        statement  = paste0(
          as.character(key$CellTypeID),
          " Travel Consistency: ", dir_txt,
          " (Welch t-test, p = ", signif(p, 3), ", ", star_from_p(p), ")"
        )
      )
    }) %>%
    ungroup()
}

ci_pairwise <- pairwise_ci_den(CI_df)
ci_pairwise
ci_pairwise$statement
```

#### Viz
```{r}
fill_cols  <- group_colors
point_cols <- point_colors

sig_rows <- ci_pairwise %>%
  filter(!is.na(sig), sig != "ns") %>%
  mutate(group1 = "Den1", group2 = "Den10")

data_max <- max(CI_df$CI, na.rm = TRUE)
if (!is.finite(data_max)) stop("CI is all NA/non-finite. Check TrackLength + displacement calculations.")

# CI should be in [0, 1] but give headroom for brackets
base_y <- min(0.92, data_max * 1.10)
step_y <- 0.06

stat_df_ci <- sig_rows %>%
  group_by(CellTypeID) %>%
  mutate(
    y.position = base_y + step_y * (row_number() - 1),
    label      = sig
  ) %>%
  ungroup() %>%
  select(CellTypeID, group1, group2, y.position, label)

ylim_top_ci <- if (nrow(stat_df_ci) > 0) {
  min(1.05, max(stat_df_ci$y.position, na.rm = TRUE) + 0.06)
} else {
  min(1.05, max(0.85, data_max * 1.10))
}

y_breaks_ci <- seq(0, 1.2, by = 0.2)

p_ci <- ggplot(CI_df, aes(x = Group, y = CI)) +
  geom_violin(
    aes(fill = Group),
    trim = FALSE,
    alpha = 0.35,
    color = NA,
    width = 0.95
  ) +
  geom_boxplot(
    aes(fill = Group),
    width = 0.22,
    outlier.shape = NA,
    alpha = 0.85,
    linewidth = 0.8
  ) +
  geom_jitter(
    aes(color = Group),
    width = 0.15,
    size = 2,
    alpha = 0.55,
    show.legend = FALSE
  ) +
  facet_grid(. ~ CellTypeID) +
  scale_fill_manual(values = fill_cols, drop = FALSE) +
  scale_color_manual(values = point_cols, drop = FALSE) +
  scale_x_discrete(labels = c("Den1" = "Low Den", "Den10" = "High Den")) +
  scale_y_continuous(breaks = y_breaks_ci, expand = c(0, 0)) +
  coord_cartesian(ylim = c(0, 1.2)) +
  {
    if (nrow(stat_df_ci) > 0) {
      stat_pvalue_manual(
        stat_df_ci,
        xmin = "group1",
        xmax = "group2",
        label = "label",
        y.position = "y.position",
        tip.length   = 0.03,
        size         = 6,
        bracket.size = 0.9
      )
    } else NULL
  } +
  labs(
    title = NULL,
    x     = "Cancer Density",
    y     = "Travel Consistency \n(Net / Total Distance)"
  ) +
  theme_classic(base_size = 20) +
  theme(
    axis.line    = element_line(linewidth = 0.9, color = "black"),
    axis.text.y  = element_text(size = 18),
    axis.title.y = element_text(size = 20),
    axis.text.x  = element_text(size = 16, angle = 0, hjust = 0.5),
    axis.title.x = element_text(size = 16),
    legend.position = "none",
    strip.background = element_blank(),
    strip.text.x = element_text(size = 20),
    panel.spacing.x = unit(2, "lines")
  )

ggsave(
  filename = "TravelConsistency_ViolinBox_StarsOnly_Den1_Den10.png",
  plot     = p_ci,
  width    = 6.5,
  height   = 4.7,
  dpi      = 600
)

p_ci
```

### 8.) Nearest-neighbor distance over time MDA -> Myoep
```{r}
library(RANN)

frame_interval_min <- 10
hours_per_frame <- frame_interval_min / 60

CP_clean <- Combined_Positions %>%
  dplyr::mutate(
    Time_num = suppressWarnings(readr::parse_number(as.character(Time))),
    X        = suppressWarnings(as.numeric(X_Position)),
    Y        = suppressWarnings(as.numeric(Y_Position))
  ) %>%
  dplyr::filter(
    is.finite(Time_num), is.finite(X), is.finite(Y),
    Group %in% c("Den1","Den10"),
    CellTypeID %in% c("MDAMB231","Myoepithelial")
  )

compute_nn_mda_to_myo <- function(df_clean) {

  frames <- df_clean %>%
    dplyr::group_by(SampleID, Group, Time_num) %>%
    dplyr::group_split()

  purrr::map_dfr(frames, function(df_t) {

    sample_id <- df_t$SampleID[1]
    group     <- df_t$Group[1]
    tnum      <- df_t$Time_num[1]

    mda <- df_t %>% dplyr::filter(CellTypeID == "MDAMB231")
    myo <- df_t %>% dplyr::filter(CellTypeID == "Myoepithelial")

    if (nrow(mda) == 0 || nrow(myo) == 0) return(NULL)

    mda_mat <- as.matrix(mda[, c("X","Y")])
    myo_mat <- as.matrix(myo[, c("X","Y")])

    nn <- RANN::nn2(data = myo_mat, query = mda_mat, k = 1)

    idx <- nn$nn.idx[, 1]
    d1  <- nn$nn.dists[, 1]

    neighbor <- myo[idx, , drop = FALSE]

    tibble::tibble(
      SampleID       = sample_id,
      Group          = group,
      Time_frame     = tnum,
      Time_hr        = tnum * hours_per_frame,
      MDA_UniqueID   = mda$UniqueID,
      MDA_X          = mda$X,
      MDA_Y          = mda$Y,
      NN_Dist_um     = d1,
      NN_UniqueID    = neighbor$UniqueID,
      NN_CellTypeID  = neighbor$CellTypeID
    )
  })
}

nn_mda_to_myo <- compute_nn_mda_to_myo(CP_clean)

# quick sanity
nn_mda_to_myo %>% dplyr::count(Group) %>% print()
nn_mda_to_myo %>% dplyr::summarise(min_d = min(NN_Dist_um), max_d = max(NN_Dist_um)) %>% print()
```

#### Viz
```{r}
nn_time_summary <- nn_mda_to_myo %>%
  dplyr::group_by(Group, Time_hr) %>%
  dplyr::summarise(
    mean_dist = mean(NN_Dist_um, na.rm = TRUE),
    sem_dist  = sd(NN_Dist_um, na.rm = TRUE) / sqrt(sum(is.finite(NN_Dist_um))),
    n_cells   = sum(is.finite(NN_Dist_um)),
    .groups   = "drop"
  ) %>%
  dplyr::mutate(Group = factor(Group, levels = c("Den1","Den10")))

# Optional: cap x-axis to your experiment length
xmax_hr <- max(nn_time_summary$Time_hr, na.rm = TRUE)
x_breaks <- seq(0, ceiling(xmax_hr / 6) * 6, by = 6)

p_nn_mean <- ggplot(nn_time_summary,
                    aes(x = Time_hr, y = mean_dist, color = Group, fill = Group)) +
  geom_ribbon(
    aes(ymin = pmax(mean_dist - sem_dist, 0),
        ymax = mean_dist + sem_dist),
        alpha = 0.18, 
        color = NA) +
  geom_line(linewidth = 1.0) +
  scale_color_manual(
  values = group_colors,
  labels = c("Den1" = "Low Den", "Den10" = "High Den"),
  drop = FALSE
) +
  scale_fill_manual(
  values = group_colors,
  labels = c("Den1" = "Low Den", "Den10" = "High Den"),
  drop = FALSE
) +
  scale_x_continuous(limits = c(0,xmax_hr), breaks = x_breaks, expand = c(0, 0)) +
  scale_y_continuous(limits = c(0,160)) +
  labs(
    x = "Time (hours)",
    y = "Mean Distance to Nearest \nMyoep Neighbor (µm)",
    title = "MDAMB231 to Nearest Myoepithelial"
  ) +
  theme_classic(base_size = 20) +
  theme(
    legend.position = "right",
    legend.title = element_blank(),
    legend.text  = element_text(size = 16),
    axis.line    = element_line(linewidth = 0.9, color = "black"),
    axis.text    = element_text(size = 18),
    axis.title   = element_text(size = 20),
    strip.background = element_blank()
  )

ggsave("NN_Distance_MDAMB231_to_Myoep_mean_over_time_Den1_Den10.png",
       p_nn_mean, width = 8.5, height = 4.8, dpi = 600)

p_nn_mean
```

### 9.) Center of Mass (COM) displacement — MDAMB231 and Myoepithelial
```{r}
frame_interval_min <- 10
hours_per_frame <- frame_interval_min / 60

star_from_p <- function(p) {
  if (is.na(p)) return(NA_character_)
  if (p <= 0.0001) return("****")
  if (p <= 0.001)  return("***")
  if (p <= 0.01)   return("**")
  if (p <= 0.05)   return("*")
  "ns"
}

prep_positions_for_com <- function(dat, celltype_use) {
  dat %>%
    mutate(
      frame = suppressWarnings(readr::parse_number(as.character(Time))) %>% as.integer(),
      X     = suppressWarnings(as.numeric(X_Position)),
      Y     = suppressWarnings(as.numeric(Y_Position))
    ) %>%
    filter(
      CellTypeID == celltype_use,
      Group %in% c("Den1","Den10"),
      is.finite(frame), is.finite(X), is.finite(Y)
    ) %>%
    arrange(SampleID, Group, UniqueID, frame) %>%
    distinct(SampleID, UniqueID, frame, .keep_all = TRUE) %>%
    mutate(Time_hours = frame * frame_interval_min / 60)
}

compute_com_centered <- function(cp_clean) {
  # COM per sample per frame
  com_raw <- cp_clean %>%
    group_by(SampleID, Group, frame) %>%
    summarise(
      com_x = mean(X, na.rm = TRUE),
      com_y = mean(Y, na.rm = TRUE),
      Time_hours = first(Time_hours),
      .groups = "drop"
    )

  # re-center each sample so t0 = (0,0)
  com_centered <- com_raw %>%
    group_by(SampleID, Group) %>%
    arrange(Time_hours, .by_group = TRUE) %>%
    mutate(
      com_x0 = first(com_x[is.finite(com_x)]),
      com_y0 = first(com_y[is.finite(com_y)]),
      dx     = com_x - com_x0,
      dy     = com_y - com_y0
    ) %>%
    ungroup()

  com_centered
}

summarise_com_time <- function(com_centered) {
  com_centered %>%
    group_by(Group, Time_hours) %>%
    summarise(
      n_repl_dx = sum(is.finite(dx)),
      n_repl_dy = sum(is.finite(dy)),
      dx_mean   = mean(dx, na.rm = TRUE),
      dy_mean   = mean(dy, na.rm = TRUE),
      dx_sem    = if_else(n_repl_dx >= 2, sd(dx, na.rm = TRUE) / sqrt(n_repl_dx), NA_real_),
      dy_sem    = if_else(n_repl_dy >= 2, sd(dy, na.rm = TRUE) / sqrt(n_repl_dy), NA_real_),
      .groups = "drop"
    ) %>%
    mutate(Group = factor(Group, levels = c("Den1","Den10")))
}

plot_COM_axis <- function(df, axis_label, y_limits = NULL, y_breaks = NULL) {
  p <- ggplot(df, aes(x = Time_hours, y = mean, color = Group, fill = Group)) +
    geom_hline(yintercept = 0, linewidth = 0.6, color = "grey70") +
    geom_ribbon(aes(ymin = mean - sem, ymax = mean + sem), alpha = 0.18, color = NA) +
    geom_line(linewidth = 1.1, na.rm = TRUE) +
    scale_color_manual(values = group_colors, drop = FALSE) +
    scale_fill_manual(values  = group_colors, drop = FALSE) +
    labs(
      x = "Time (hours)",
      y = axis_label,
      color = "Cancer\nDensity",
      fill  = "Cancer\nDensity"
    ) +
    theme_classic(base_size = 20) +
    theme(
      legend.position = "right",
      legend.title = element_text(size = 16),
      legend.text  = element_text(size = 14),
      axis.line    = element_line(linewidth = 0.9, color = "black"),
      axis.text    = element_text(size = 16),
      axis.title   = element_text(size = 18)
    )

  if (!is.null(y_limits)) p <- p + scale_y_continuous(limits = y_limits, breaks = y_breaks)
  p
}

compute_end_vectors_by_repl <- function(com_centered, tmax_hr = NULL) {
  d <- com_centered %>%
    filter(is.finite(Time_hours), is.finite(dx), is.finite(dy))

  if (!is.null(tmax_hr)) d <- d %>% filter(Time_hours <= tmax_hr)

  d %>%
    group_by(Group, SampleID) %>%
    summarise(
      Time_end = max(Time_hours, na.rm = TRUE),
      dx_end   = dx[which.max(Time_hours)],
      dy_end   = dy[which.max(Time_hours)],
      bearing_deg = atan2(dy_end, dx_end) * 180 / pi,
      .groups = "drop"
    ) %>%
    mutate(Group = factor(Group, levels = c("Den1","Den10")))
}

plot_end_vectors <- function(repl_df, axis_lim = NULL, title = NULL) {
  p <- ggplot(repl_df, aes(color = Group)) +
    geom_hline(yintercept = 0, linewidth = 0.4, color = "grey85") +
    geom_vline(xintercept = 0, linewidth = 0.4, color = "grey85") +
    geom_segment(
      aes(x = 0, y = 0, xend = dx_end, yend = dy_end),
      arrow = arrow(length = unit(2.2, "mm"), type = "closed"),
      linewidth = 1.0,
      alpha = 0.95
    ) +
    scale_color_manual(values = group_colors, drop = FALSE) +
    coord_fixed(expand = FALSE) +
    labs(
      title = title,
      x = "ΔX COM displacement (µm)",
      y = "ΔY COM displacement (µm)",
      color = "Cancer\nDensity"
    ) +
    theme_classic(base_size = 18) +
    theme(
      legend.position = "right",
      legend.title = element_text(size = 14),
      legend.text  = element_text(size = 12),
      axis.text    = element_text(size = 14),
      axis.title   = element_text(size = 16)
    )

  if (!is.null(axis_lim)) {
    p <- p + coord_fixed(xlim = c(-axis_lim, axis_lim), ylim = c(-axis_lim, axis_lim), expand = FALSE)
  }
  p
}

# time-averaged COM per replicate, then stats + boxplot
compute_timeavg_com_by_repl <- function(com_centered) {
  com_centered %>%
    filter(is.finite(dx), is.finite(dy)) %>%
    group_by(SampleID, Group) %>%
    summarise(
      dX_mean = mean(dx, na.rm = TRUE),
      dY_mean = mean(dy, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(Group = factor(Group, levels = c("Den1","Den10"))) %>%
    pivot_longer(cols = c(dX_mean, dY_mean),
                 names_to = "Component",
                 values_to = "Value") %>%
    mutate(
      Component = recode(Component,
                         dX_mean = "ΔX (µm)",
                         dY_mean = "ΔY (µm)"),
      Component = factor(Component, levels = c("ΔX (µm)", "ΔY (µm)"))
    )
}

welch_box_stats_by_component <- function(df_box) {

  res <- df_box %>%
    group_by(Component) %>%
    group_modify(function(d, key) {

      d <- d %>% tidyr::drop_na(Value, Group)

      # require both groups + >=2 replicates per group
      if (n_distinct(d$Group) < 2 || any(table(d$Group) < 2)) {
        return(tibble(
          p_value    = NA_real_,
          direction  = NA_character_,
          sig        = NA_character_,
          statement  = paste0(as.character(key$Component), ": Not enough replicates in one group")
        ))
      }

      tt <- t.test(Value ~ Group, data = d, var.equal = FALSE)
      p  <- tt$p.value

      means <- tapply(d$Value, d$Group, mean, na.rm = TRUE)
      dir_txt <- if (means["Den1"] > means["Den10"]) "Den1 > Den10" else "Den10 > Den1"

      tibble(
        p_value   = p,
        direction = dir_txt,
        sig       = star_from_p(p),
        statement = paste0(
          as.character(key$Component), ": ", dir_txt,
          " (Welch t-test p = ", signif(p, 3), ", ", star_from_p(p), ")"
        )
      )
    }) %>%
    ungroup()

  res
}

build_stat_df_for_box <- function(df_box, stats_df) {
  # Stars only, one bracket per component (if significant by raw p)
  out <- stats_df %>%
    filter(is.finite(p_value), sig != "ns") %>%
    mutate(
      group1 = "Den1",
      group2 = "Den10"
    )

  if (nrow(out) == 0) return(tibble())

  out %>%
    group_by(Component) %>%
    mutate(
      ymax_comp = max(df_box$Value[df_box$Component == Component[1]], na.rm = TRUE),
      ymin_comp = min(df_box$Value[df_box$Component == Component[1]], na.rm = TRUE),
      step = (ymax_comp - ymin_comp) * 0.12,
      step = ifelse(is.finite(step) & step > 0, step, 0.1),
      y.position = ymax_comp * 1.10 + step * (row_number() - 1L),
      label = sig
    ) %>%
    ungroup() %>%
    select(Component, group1, group2, y.position, label)
}

plot_timeavg_box <- function(df_box, pval_df, title = NULL) {
  y_min <- min(df_box$Value, na.rm = TRUE)
  y_max_data <- max(df_box$Value, na.rm = TRUE)
  y_max <- if (nrow(pval_df) > 0) max(y_max_data, max(pval_df$y.position, na.rm = TRUE)) * 1.05 else y_max_data * 1.10

  p <- ggplot(df_box, aes(x = Group, y = Value, fill = Group)) +
    geom_hline(yintercept = 0, linewidth = 0.6, color = "grey55") +
    geom_boxplot(width = 0.6, alpha = 0.9, color = "black", outlier.shape = NA) +
    geom_point(position = position_jitter(width = 0.25, height = 0),
               size = 2.5, alpha = 0.9, color = "black") +
    facet_grid(~ Component, switch = "x") +
    scale_x_discrete(labels = c("Den1" = "Low Den", "Den10" = "High Den")) +
    scale_fill_manual(values = group_colors, drop = FALSE) +
    coord_cartesian(ylim = c(y_min * 1.05, y_max)) +
    labs(title = title, x = NULL, y = "Time-averaged COM displacement (µm)") +
    theme_classic(base_size = 18) +
    theme(
      legend.position  = "none",
      strip.placement  = "outside",
      strip.background = element_blank(),
      strip.text       = element_text(size = 16),
      axis.text        = element_text(size = 14),
      axis.title       = element_text(size = 16),
      panel.spacing    = unit(2, "lines")
    )

  if (nrow(pval_df) > 0) {
    p <- p + ggpubr::stat_pvalue_manual(
      data = pval_df,
      label = "label",
      xmin = "group1",
      xmax = "group2",
      y.position = "y.position",
      tip.length = 0.02,
      size = 7,
      bracket.size = 0.7
    )
  }
  p
}

run_COM_for_celltype <- function(celltype_use,
                                 y_limits = NULL,
                                 y_breaks = NULL,
                                 axis_lim_vec = NULL,
                                 tmax_hr_vec = NULL) {

  cp_clean <- prep_positions_for_com(Combined_Positions, celltype_use)
  if (nrow(cp_clean) == 0) stop(paste0("No data found for CellTypeID = ", celltype_use))

  com_centered <- compute_com_centered(cp_clean)
  com_summary  <- summarise_com_time(com_centered)

  # ΔX and ΔY time-series
  dx_df <- com_summary %>% transmute(Group, Time_hours, mean = dx_mean, sem = dx_sem)
  dy_df <- com_summary %>% transmute(Group, Time_hours, mean = dy_mean, sem = dy_sem)

  p_dx <- plot_COM_axis(dx_df, axis_label = "ΔX Center of Mass Displacement (µm)", y_limits = y_limits, y_breaks = y_breaks)
  p_dy <- plot_COM_axis(dy_df, axis_label = "ΔY Center of Mass Displacement (µm)", y_limits = y_limits, y_breaks = y_breaks)

  ggsave(paste0(celltype_use, "_COM_Displacement_DX.png"), p_dx, width = 7.5, height = 4.8, dpi = 600)
  ggsave(paste0(celltype_use, "_COM_Displacement_DY.png"), p_dy, width = 7.5, height = 4.8, dpi = 600)

  # End vectors by replicate
  repl_vec <- compute_end_vectors_by_repl(com_centered, tmax_hr = tmax_hr_vec)

  p_vec <- plot_end_vectors(
    repl_vec,
    axis_lim = axis_lim_vec,
    title = paste0(celltype_use, " — COM End Vectors (by replicate)")
  )
  ggsave(paste0(celltype_use, "_COM_EndVectors_ByReplicate.png"), p_vec, width = 5.2, height = 5.2, dpi = 600)

  # Time-averaged boxplots + stats (ΔX, ΔY)
  df_box <- compute_timeavg_com_by_repl(com_centered)
  stats_box <- welch_box_stats_by_component(df_box)
  pval_df <- build_stat_df_for_box(df_box, stats_box)

  p_box <- plot_timeavg_box(df_box, pval_df, title = paste0(celltype_use, " — Time-averaged COM"))
  ggsave(paste0(celltype_use, "_COM_TimeAveraged_dX_dY_boxplot_stars.png"),
         p_box, width = 7.2, height = 4.4, dpi = 600)

  list(
    cp_clean = cp_clean,
    com_centered = com_centered,
    com_summary = com_summary,
    end_vectors = repl_vec,
    timeavg_box_df = df_box,
    timeavg_stats = stats_box,
    p_dx = p_dx,
    p_dy = p_dy,
    p_vec = p_vec,
    p_box = p_box
  )
}


# MDAMB231
COM_MDA <- run_COM_for_celltype(
  celltype_use = "MDAMB231",
  y_limits = NULL,
  y_breaks = NULL,
  axis_lim_vec = NULL,
  tmax_hr_vec = NULL
)

# Myoepithelial
COM_Myo <- run_COM_for_celltype(
  celltype_use = "Myoepithelial",
  y_limits = NULL,
  y_breaks = NULL,
  axis_lim_vec = NULL,
  tmax_hr_vec = NULL
)

# Inspect stats text quickly
COM_MDA$timeavg_stats$statement
COM_Myo$timeavg_stats$statement

# Display plots in-session
COM_MDA$p_dx; COM_MDA$p_dy; COM_MDA$p_vec; COM_MDA$p_box
COM_Myo$p_dx; COM_Myo$p_dy; COM_Myo$p_vec; COM_Myo$p_box

```


# --------------------------------------------------------------------------------------------------
# XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX
# --------------------------------------------------------------------------------------------------

# Cancer Organization
```{r}
celltype_colors <- c(
  "MDAMB231"     = "#B64249",
  "Myoepithelial"= "#796089",
  "Missing_ID"   = "#AFABAB")

group_colors <- c("Segregated" = "#92A594", "Clustered" = "#C1D6E2")

point_colors <- c("Segregated" = "#6C816E", "Clustered" = "#AAC2CD")
```


## Load data & preprocess
```{r}
folder <- "/Users/helmsh/Desktop/" 

files <- c(
  "Myoep_MDAMB231_Seg1.xlsx",
  "Myoep_MDAMB231_Seg2.xlsx",
  "Myoep_MDAMB231_Seg3.xlsx",
  "Myoep_MDAMB231_Clus1.xlsx",
  "Myoep_MDAMB231_Clus2.xlsx",
  "Myoep_MDAMB231_Clus3.xlsx")

# SampleID = Seg1, Seg2, ..., Clus3
sample_names <- files %>%
  str_remove("Myoep_MDAMB231_") %>%
  str_remove("\\.xlsx$")

# Group = Seg or Clus
group_names <- str_extract(sample_names, "Seg|Clus") %>%
  recode(
    "Seg"  = "Segregated",
    "Clus" = "Clustered"
  )

unique(sample_names)
unique(group_names)

# Load files
data_list <- lapply(seq_along(files), function(i) {
  read_excel(
    path = file.path(folder, files[i]),
    sheet = "Position",
    skip = 1,
    col_types = "text"
  ) %>%
    rename(CellTypeID = 7) %>%  
    mutate(
      SampleID = sample_names[i],
      Group    = group_names[i]
    )
})

# Combine into one dataframe
all_positions <- bind_rows(data_list)

# Clean CellTypeID to exactly 2 levels
all_positions <- all_positions %>%
  mutate(
    CellTypeID = case_when(
      grepl("^Myo", CellTypeID, ignore.case = TRUE) ~ "Myoepithelial",
      grepl("^MDA", CellTypeID, ignore.case = TRUE) ~ "MDAMB231",
      grepl("231",  CellTypeID, ignore.case = TRUE) ~ "MDAMB231",
      TRUE ~ CellTypeID
    )
  )

# sanity checks
unique(all_positions$CellTypeID)
unique(all_positions$Group)
unique(all_positions$SampleID)
head(all_positions)
```

```{r}
# Add Unique ID - Imaris reuses track ID's across files
all_positions <- all_positions %>%
  mutate(UniqueID = paste(SampleID, TrackID, sep = "_"))

head(all_positions)
```

```{r}
# Confirm everything has a uniqueID. These datasets have 145 frames. If there are more than that it means there is a duplicate
all_positions %>%
  count(UniqueID) %>%
  arrange(desc(n))
```

```{r}
# Can also confirm UniqueID through this. If any UniqueID maps to >1 SampleID or >1 TrackID, something is wrong
uid_integrity <- all_positions %>%
  group_by(UniqueID) %>%
  summarise(
    n_sample = n_distinct(SampleID),
    n_track  = n_distinct(TrackID),
    n_group  = n_distinct(Group),
    .groups = "drop"
  ) %>%
  filter(n_sample > 1 | n_track > 1 | n_group > 1)

uid_integrity
```

```{r}
missing_tracks <- all_positions %>%
  filter(CellTypeID == "Missing_ID") %>%
  distinct(SampleID, UniqueID)

missing_tracks
n_missing <- nrow(missing_tracks)
n_missing
```

```{r}
# Remove spaces in col names, i.e. "Position X" to "X_Position". Spaces cause problems. Also renaming ID because it seemed to want to pull functions instead of data
all_positions <- all_positions %>%
  rename(
    X_Position = `Position X`,
    Y_Position = `Position Y`,
    SpotID     = `ID`)

head(all_positions)
```

```{r}
# Make SpotID all the same number of characters for easier sorting
all_positions$SpotID <- str_pad(all_positions$SpotID, width = 6, side = "left", pad = "0")

head(all_positions)
```

```{r}
# One of the short track ID's is 000000. Rename to 999999
all_positions$SpotID[all_positions$SpotID == "000000"] <- "999999"

head(all_positions)
```

```{r}
# Make time all the same number of characters
all_positions$Time <- str_pad(all_positions$Time, width = 3, side = "left", pad = "0")

head(all_positions)
```

```{r}
# Reorder cols so it is easier to work with
all_positions_cleaned <- all_positions %>%
  select(SampleID, UniqueID, CellTypeID, Time, X_Position, Y_Position, Group, Unit, Category, Collection, TrackID, SpotID)

head(all_positions_cleaned)
```

```{r}
# Rearrange so that is sorted by SampleID, then UniqueID, then Time so we can separate out the duplicates
Combined_Positions <- all_positions_cleaned %>% arrange(SampleID, UniqueID, Time)

head(Combined_Positions)
```

```{r}
unique(Combined_Positions$CellTypeID)
unique(Combined_Positions$Group)
unique(Combined_Positions$SampleID)
```


## Analysis
### 1.) Migration Track Plots
#### Raw Data - Each Sample Individually
```{r}
samples_to_plot <- sort(unique(Combined_Positions$SampleID))

span_x <- 800
span_y <- 800
tick   <- 200

for (sample_to_plot in samples_to_plot) {

  df_tracks <- Combined_Positions %>%
    filter(SampleID == sample_to_plot, CellTypeID != "Missing_ID") %>%
    mutate(
      TimeNum = suppressWarnings(readr::parse_number(Time)),
      X       = suppressWarnings(readr::parse_number(X_Position)),
      Y       = suppressWarnings(readr::parse_number(Y_Position))
    ) %>%
    filter(is.finite(TimeNum), is.finite(X), is.finite(Y)) %>%
    arrange(UniqueID, TimeNum)

  if (nrow(df_tracks) == 0) next

  # Define the viewing window using the start frame (Time == "001")
  df_start <- df_tracks %>% filter(Time == "001")
  if (nrow(df_start) == 0) next

  # Center of the start frame bounding box
  x0 <- mean(range(df_start$X, na.rm = TRUE))
  y0 <- mean(range(df_start$Y, na.rm = TRUE))

  # Start with a centered window
  x_centered_limits <- c(x0 - span_x/2, x0 + span_x/2)
  y_centered_limits <- c(y0 - span_y/2, y0 + span_y/2)

  # but convert it to "xmin anchored" integer limits with no decimals:
  x_min <- as.integer(floor(x_centered_limits[1]))
  y_min <- as.integer(floor(y_centered_limits[1]))

  x_limits <- c(x_min, x_min + span_x)
  y_limits <- c(y_min, y_min + span_y)

  # Ticks: xmin/ymin + 200 increments, total span 800
  x_breaks <- seq(x_limits[1], x_limits[2], by = tick)
  y_breaks <- seq(y_limits[1], y_limits[2], by = tick)

  p_tracks <- ggplot(
    df_tracks,
    aes(x = X, y = Y, group = UniqueID, color = CellTypeID)
  ) +
    geom_path(linewidth = 0.6, alpha = 0.6, na.rm = TRUE) +
    scale_x_continuous(limits = x_limits, breaks = x_breaks, expand = c(0, 0)) +
    scale_y_continuous(limits = y_limits, breaks = y_breaks, expand = c(0, 0)) +
    scale_color_manual(values = celltype_colors, drop = FALSE) +
    labs(
      title = paste0(sample_to_plot, " — All tracks"),
      x = "X (µm)",
      y = "Y (µm)",
      color = "Cell type"
    ) +
    theme_classic(base_size = 20) +
    theme(
      axis.line = element_line(linewidth = 0.6),
      axis.title = element_text(size = 16),
      axis.text  = element_text(size = 12),
      plot.title = element_text(size = 20, face = "bold")
    )

  ggsave(
    filename = paste0(sample_to_plot, "_AllTracks_FIXEDWINDOW.png"),
    plot = p_tracks,
    width = 6,
    height = 4,
    dpi = 600
  )

  print(p_tracks)
}
```

#### Centered
```{r}
# Find bounding box to align the samples
bb_by_sample <- Combined_Positions %>%
  mutate(
    X = suppressWarnings(readr::parse_number(X_Position)),
    Y = suppressWarnings(readr::parse_number(Y_Position))
  ) %>%
  filter(is.finite(X), is.finite(Y)) %>%
  group_by(SampleID) %>%
  summarise(
    xmin = min(X, na.rm = TRUE),
    xmax = max(X, na.rm = TRUE),
    ymin = min(Y, na.rm = TRUE),
    ymax = max(Y, na.rm = TRUE),
    x_range = xmax - xmin,
    y_range = ymax - ymin,
    .groups = "drop"
  ) %>%
  arrange(SampleID)

bb_by_sample

bb_by_sample %>%
  summarise(
    max_x_range = max(x_range, na.rm = TRUE),
    max_y_range = max(y_range, na.rm = TRUE)
  )
```

##### Individual samples
```{r}
# Center each SampleID using only the start frame (Time == "001"). Then apply that shift to all timepoints in that sample

x_limits <- c(-451.85, 451.85)
y_limits <- c(-453.06, 454.06)
axis_breaks <- c(-450, -225, 0, 225, 450)

CP_start_centered <- Combined_Positions %>%
  mutate(
    TimeNum = suppressWarnings(readr::parse_number(Time)),
    X = suppressWarnings(readr::parse_number(X_Position)),
    Y = suppressWarnings(readr::parse_number(Y_Position))
  ) %>%
  filter(is.finite(TimeNum), is.finite(X), is.finite(Y)) %>%

  # compute per-sample start-frame center (Time == "001")
  group_by(SampleID) %>%
  mutate(
    start_x_mid = {
      xs <- X[Time == "001"]
      (min(xs, na.rm = TRUE) + max(xs, na.rm = TRUE)) / 2
    },
    start_y_mid = {
      ys <- Y[Time == "001"]
      (min(ys, na.rm = TRUE) + max(ys, na.rm = TRUE)) / 2
    },

    # shift all timepoints by that start-frame center
    Xc = X - start_x_mid,
    Yc = Y - start_y_mid
  ) %>%
  ungroup()

plot_tracks_fixedbox_startcenter <- function(sample_to_plot, cell_filter = c("all", "MDAMB231", "Myoepithelial")) {

  cell_filter <- match.arg(cell_filter)

  df_tracks <- CP_start_centered %>%
    filter(SampleID == sample_to_plot, CellTypeID != "Missing_ID") %>%
    { if (cell_filter == "all") . else filter(., CellTypeID == cell_filter) } %>%
    arrange(UniqueID, TimeNum)

  # if a sample is missing that cell type, just return NULL
  if (nrow(df_tracks) == 0) return(NULL)

  ggplot(df_tracks, aes(x = Xc, y = Yc, group = UniqueID, color = CellTypeID)) +
    geom_path(linewidth = 0.6, alpha = 0.6, na.rm = TRUE) +
    coord_equal() +
    scale_x_continuous(limits = x_limits, breaks = axis_breaks, expand = c(0, 0)) +
    scale_y_continuous(limits = y_limits, breaks = axis_breaks, expand = c(0, 0)) +
    scale_color_manual(values = celltype_colors, drop = FALSE) +
    labs(
      title = paste0(sample_to_plot, " centered by start frame (Time 001) — ", cell_filter),
      x = "X (µm)",
      y = "Y (µm)",
      color = "Cell type"
    ) +
  theme_classic(base_size = 20) +
  theme(
    axis.line    = element_line(linewidth = 0.9, color = "black"),
    axis.text.y  = element_text(size = 18),
    axis.title.y = element_text(size = 20),
    axis.text.x  = element_text(size = 16, angle = 0, hjust = 0.5),
    axis.title.x = element_text(size = 16),
    legend.position = "none",
    strip.background = element_blank(),
    strip.text.x = element_text(size = 20),
    panel.spacing.x = unit(2, "lines")
  )
}

## Loop and save all samples: All tracks + MDAMB231-only + Myoep-only
samples <- c("Clus1","Clus2","Clus3","Seg1","Seg2","Seg3")

for (sid in samples) {

  # 1) all tracks
  p_all <- plot_tracks_fixedbox_startcenter(sid, "all")
  if (!is.null(p_all)) {
    ggsave(
      filename = paste0(sid, "_Tracks_All_StartCentered_FixedBox.png"),
      plot = p_all,
      width = 7,
      height = 6,
      dpi = 600
    )
  }

  # 2) MDAMB231-only
  p_mda <- plot_tracks_fixedbox_startcenter(sid, "MDAMB231")
  if (!is.null(p_mda)) {
    ggsave(
      filename = paste0(sid, "_Tracks_MDAMB231_StartCentered_FixedBox.png"),
      plot = p_mda,
      width = 7,
      height = 6,
      dpi = 600
    )
  }

  # 3) Myoep-only
  p_myo <- plot_tracks_fixedbox_startcenter(sid, "Myoepithelial")
  if (!is.null(p_myo)) {
    ggsave(
      filename = paste0(sid, "_Tracks_Myoepithelial_StartCentered_FixedBox.png"),
      plot = p_myo,
      width = 7,
      height = 6,
      dpi = 600
    )
  }
}
```

##### Grouped
```{r}
x_limits <- c(-451.85, 451.85)
y_limits <- c(-453.06, 454.06)
axis_breaks <- c(-450, -225, 0, 225, 450)

plot_group_tracks_fixedbox_startcenter <- function(group_to_plot,
                                                  cell_filter = c("all", "MDAMB231", "Myoepithelial")) {

  cell_filter <- match.arg(cell_filter)

  df_tracks <- CP_start_centered %>%
    filter(Group == group_to_plot, CellTypeID != "Missing_ID") %>%
    { if (cell_filter == "all") . else dplyr::filter(., CellTypeID == cell_filter) } %>%
    arrange(SampleID, UniqueID, TimeNum)

  if (nrow(df_tracks) == 0) return(NULL)

  ggplot(
    df_tracks,
    aes(
      x = Xc,
      y = Yc,
      group = UniqueID,        
      color = CellTypeID
    )
  ) +
    geom_path(linewidth = 0.6, alpha = 0.6, na.rm = TRUE) +
    coord_equal() +
    scale_x_continuous(limits = x_limits, breaks = axis_breaks, expand = c(0, 0)) +
    scale_y_continuous(limits = y_limits, breaks = axis_breaks, expand = c(0, 0)) +
    scale_color_manual(values = celltype_colors, drop = FALSE) +
    theme_classic(base_size = 20) +
    labs(
      title = paste0(group_to_plot, cell_filter),
      x = "X (µm)",
      y = "Y (µm)",
      color = "Cell type"
    )
}

## Make + save all 6 plot variations:
groups_to_plot <- c("Clustered", "Segregated")
cell_filters <- c("all", "MDAMB231", "Myoepithelial")

for (g in groups_to_plot) {
  for (cf in cell_filters) {

    p <- plot_group_tracks_fixedbox_startcenter(g, cf)

    if (!is.null(p)) {
      ggsave(
        filename = paste0(g, "_Tracks_", cf, "_Centered.png"),
        plot = p,
        width = 7,
        height = 6,
        dpi = 600
      )
    }
  }
}
```

#### Relative Position
```{r}
#### Relative Position (Spatial dataset: Segregated vs Clustered)

# 1) Build relative coordinates per track (anchor at first observed non-NA position)
Combined_Positions_Relative <- Combined_Positions %>%
  mutate(
    Xnum = suppressWarnings(readr::parse_number(as.character(X_Position))),
    Ynum = suppressWarnings(readr::parse_number(as.character(Y_Position))),
    Tnum = suppressWarnings(readr::parse_number(as.character(Time))) %>% as.integer()
  ) %>%
  filter(
    is.finite(Tnum), is.finite(Xnum), is.finite(Ynum),
    CellTypeID != "Missing_ID",
    Group %in% c("Segregated", "Clustered"),
    CellTypeID %in% c("MDAMB231", "Myoepithelial")
  ) %>%
  arrange(SampleID, UniqueID, Tnum) %>%
  group_by(SampleID, UniqueID) %>%
  # anchor each track at its first *observed* (non-NA) position
  mutate(
    .x0 = {
      ix <- which(!is.na(Xnum))
      if (length(ix)) Xnum[ix[1]] else NA_real_
    },
    .y0 = {
      iy <- which(!is.na(Ynum))
      if (length(iy)) Ynum[iy[1]] else NA_real_
    },
    RelX = Xnum - .x0,
    RelY = Ynum - .y0
  ) %>%
  ungroup() %>%
  select(
    SampleID, UniqueID, CellTypeID, Time,
    X_Position, Y_Position, Group, Unit, Category, Collection, SpotID,
    RelX, RelY
  )

# Sanity check: first point of every track should be (0,0)
Combined_Positions_Relative %>%
  group_by(SampleID, UniqueID) %>%
  slice(1) %>%
  summarise(first_RelX = first(RelX), first_RelY = first(RelY), .groups = "drop") %>%
  summarise(
    n_tracks      = n(),
    n_zero_start  = sum(first_RelX == 0 & first_RelY == 0, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  print()
```

```{r}
# 2) Plot helper (fixed axes, same aesthetic as your other dataset)

CP_rel <- Combined_Positions_Relative %>%
  mutate(
    RelX = as.numeric(RelX),
    RelY = as.numeric(RelY)
  )

plot_tracks_rel_fixed <- function(df, title, show_legend = TRUE) {
  ggplot(
    df,
    aes(
      x = RelX,
      y = RelY,
      group = interaction(SampleID, UniqueID),
      color = CellTypeID
    )
  ) +
    geom_path(linewidth = 0.6, alpha = 0.6, na.rm = TRUE) +
    guides(color = guide_legend(override.aes = list(linewidth = 2))) +
    coord_equal() +
    scale_x_continuous(
      limits = c(-445, 450),
      breaks = c(-400, -200, 0, 200, 400),
      expand = c(0, 0)
    ) +
    scale_y_continuous(
      limits = c(-450, 450),
      breaks = c(-400, -200, 0, 200, 400),
      expand = c(0, 0)
    ) +
    scale_color_manual(values = celltype_colors, drop = FALSE) +
    theme_classic(base_size = 20) +
    theme(
      axis.title        = element_text(size = 20),
      axis.text         = element_text(size = 20),
      plot.title        = element_text(size = 20, face = "bold"),
      legend.title      = element_text(size = 24),
      legend.text       = element_text(size = 22),
      legend.key.height = unit(30, "pt"),
      legend.key.width  = unit(30, "pt"),
      legend.position   = if (show_legend) "right" else "none"
    ) +
    labs(
      title = title,
      x = "Relative X Position (µm)",
      y = "Relative Y Position (µm)",
      color = "Track Identity"
    )
}

seg_rel  <- CP_rel %>% filter(Group == "Segregated")
clus_rel <- CP_rel %>% filter(Group == "Clustered")
```

```{r}
# 3) Group-level exports (All / MDAMB231 / Myoepithelial)

# Segregated
p_seg_all <- plot_tracks_rel_fixed(seg_rel,  "Segregated — Relative: All Tracks", show_legend = TRUE)
p_seg_mda <- plot_tracks_rel_fixed(seg_rel %>% filter(CellTypeID == "MDAMB231"),
                                   "Segregated — Relative: MDAMB231 Tracks", show_legend = TRUE)
p_seg_myo <- plot_tracks_rel_fixed(seg_rel %>% filter(CellTypeID == "Myoepithelial"),
                                   "Segregated — Relative: Myoepithelial Tracks", show_legend = TRUE)

ggsave("Segregated_Relative_All_Tracks.png",           p_seg_all, width = 7, height = 6, dpi = 600)
ggsave("Segregated_Relative_MDAMB231_Tracks.png",      p_seg_mda, width = 7, height = 6, dpi = 600)
ggsave("Segregated_Relative_Myoepithelial_Tracks.png", p_seg_myo, width = 7, height = 6, dpi = 600)

# Clustered
p_clus_all <- plot_tracks_rel_fixed(clus_rel, "Clustered — Relative: All Tracks", show_legend = TRUE)
p_clus_mda <- plot_tracks_rel_fixed(clus_rel %>% filter(CellTypeID == "MDAMB231"),
                                    "Clustered — Relative: MDAMB231 Tracks", show_legend = TRUE)
p_clus_myo <- plot_tracks_rel_fixed(clus_rel %>% filter(CellTypeID == "Myoepithelial"),
                                    "Clustered — Relative: Myoepithelial Tracks", show_legend = TRUE)

ggsave("Clustered_Relative_All_Tracks.png",           p_clus_all, width = 7, height = 6, dpi = 600)
ggsave("Clustered_Relative_MDAMB231_Tracks.png",      p_clus_mda, width = 7, height = 6, dpi = 600)
ggsave("Clustered_Relative_Myoepithelial_Tracks.png", p_clus_myo, width = 7, height = 6, dpi = 600)

p_seg_all
p_clus_all
```

```{r}
# 4) Per-sample exports (All / MDAMB231 / Myoepithelial) with legend off

samples_rel <- sort(unique(CP_rel$SampleID))

for (s in samples_rel) {
  d <- CP_rel %>% filter(SampleID == s)

  p_all <- plot_tracks_rel_fixed(d, paste0(s, " — Relative: All Tracks"), show_legend = FALSE)
  p_mda <- plot_tracks_rel_fixed(d %>% filter(CellTypeID == "MDAMB231"),
                                 paste0(s, " — Relative: MDAMB231 Tracks"), show_legend = FALSE)
  p_myo <- plot_tracks_rel_fixed(d %>% filter(CellTypeID == "Myoepithelial"),
                                 paste0(s, " — Relative: Myoepithelial Tracks"), show_legend = FALSE)

  ggsave(paste0(s, "_Relative_All_Tracks.png"),           p_all, width = 6, height = 6, dpi = 600)
  ggsave(paste0(s, "_Relative_MDAMB231_Tracks.png"),      p_mda, width = 6, height = 6, dpi = 600)
  ggsave(paste0(s, "_Relative_Myoepithelial_Tracks.png"), p_myo, width = 6, height = 6, dpi = 600)
}
```

### 2.) Convex Hull - Migration territory 
```{r}
Combined_Positions_Relative <- Combined_Positions_Relative %>%
  dplyr::mutate(
    Group = dplyr::case_when(
      Group %in% c("Segregated", "Clustered") ~ Group,
      TRUE ~ NA_character_
    ),
    Group = factor(Group, levels = c("Segregated", "Clustered"))
  )

ConvexAreas <- Combined_Positions_Relative %>%
  dplyr::group_by(SampleID, UniqueID, CellTypeID, Group) %>%
  dplyr::summarise(
    area = {
      pts <- cbind(RelX, RelY)
      pts <- pts[complete.cases(pts), , drop = FALSE]
      if (nrow(pts) < 3) NA_real_ else geometry::convhulln(pts, "FA")$area
    },
    .groups = "drop"
  )
```

#### Hull Viz
```{r}
.clean_pts <- function(df, x = "RelX", y = "RelY") {
  pts <- df[, c(x, y), drop = FALSE]
  pts[[x]] <- as.numeric(pts[[x]])
  pts[[y]] <- as.numeric(pts[[y]])
  pts <- pts[is.finite(pts[[x]]) & is.finite(pts[[y]]), , drop = FALSE]
  pts <- dplyr::distinct(pts) # remove exact duplicates
  pts
}

hull_df <- function(df, x = "RelX", y = "RelY") {
  pts <- .clean_pts(df, x, y)
  if (nrow(pts) < 3) return(NULL)
  idx <- chull(pts[[x]], pts[[y]])
  out <- pts[idx, , drop = FALSE]
  rbind(out, out[1, , drop = FALSE]) # close polygon
}

build_per_track_hulls <- function(dat, x = "RelX", y = "RelY") {
  dat %>%
    dplyr::group_by(SampleID, UniqueID, CellTypeID) %>%
    dplyr::group_modify(~{
      h <- hull_df(.x, x, y)
      if (is.null(h)) return(tibble::tibble(RelX = numeric(0), RelY = numeric(0)))
      tibble::tibble(RelX = h[[x]], RelY = h[[y]])
    }) %>%
    dplyr::ungroup()
}

build_sample_hull <- function(dat, x = "RelX", y = "RelY") {
  h <- hull_df(dat, x, y)
  if (is.null(h)) return(NULL)
  tibble::tibble(RelX = h[[x]], RelY = h[[y]])
}

base_style <- function(show_legend = FALSE) {
  theme_classic(base_size = 20) +
    theme(
      legend.position = if (show_legend) "right" else "none",
      axis.title = element_text(size = 20),
      axis.text  = element_text(size = 20),
      plot.title = element_text(size = 20, face = "bold")
    )
}

plot_sample_with_hulls <- function(sample_id,
                                   cell_filter = c("ALL", "MDAMB231", "Myoepithelial"),
                                   per_track_hulls = TRUE,
                                   sample_hull = TRUE,
                                   fixed_limits = c(-300, 300, -300, 300),
                                   show_legend = FALSE) {

  cell_filter <- match.arg(cell_filter)

  df_plot <- Combined_Positions_Relative %>%
    dplyr::filter(SampleID == sample_id, CellTypeID != "Missing_ID") %>%
    dplyr::mutate(RelX = as.numeric(RelX), RelY = as.numeric(RelY)) %>%
    dplyr::filter(is.finite(RelX), is.finite(RelY))

  if (cell_filter != "ALL") df_plot <- df_plot %>% dplyr::filter(CellTypeID == cell_filter)
  if (!nrow(df_plot)) return(ggplot() + ggtitle(paste0(sample_id, " — EMPTY")))

  xlim <- fixed_limits[1:2]
  ylim <- fixed_limits[3:4]

  df_hull <- df_plot %>%
    dplyr::filter(
      dplyr::between(RelX, xlim[1], xlim[2]),
      dplyr::between(RelY, ylim[1], ylim[2])
    )

  per_track <- if (per_track_hulls) build_per_track_hulls(df_hull) else NULL
  smp_hull  <- if (sample_hull)      build_sample_hull(df_hull)    else NULL

  p <- ggplot(df_plot, aes(x = RelX, y = RelY, group = UniqueID, color = CellTypeID)) +
    geom_path(linewidth = 0.6, alpha = 0.6, na.rm = TRUE) +
    coord_equal() +
    scale_x_continuous(
      limits = xlim, breaks = c(-300, -150, 0, 150, 300), expand = c(0, 0)
    ) +
    scale_y_continuous(
      limits = ylim, breaks = c(-300, -150, 0, 150, 300), expand = c(0, 0)
    ) +
    scale_color_manual(values = celltype_colors, drop = FALSE) +
    scale_fill_manual(values = celltype_colors, drop = FALSE) +
    labs(
      title = paste0(sample_id, " — Convex Hulls, Relative XY (", cell_filter, ")"),
      x = "Relative X Position (µm)",
      y = "Relative Y Position (µm)",
      color = "Cell Type"
    ) +
    base_style(show_legend)

  if (!is.null(per_track) && nrow(per_track)) {
    per_track <- per_track %>%
      dplyr::left_join(
        df_hull %>% dplyr::distinct(SampleID, UniqueID, CellTypeID),
        by = c("SampleID", "UniqueID", "CellTypeID")
      )

    p <- p +
      geom_polygon(
        data = per_track,
        aes(x = RelX, y = RelY, group = UniqueID, fill = CellTypeID, color = CellTypeID),
        linewidth = 0.4, alpha = 0.15, inherit.aes = FALSE
      )
  }

  if (!is.null(smp_hull)) {
    p <- p +
      geom_path(
        data = smp_hull,
        aes(x = RelX, y = RelY),
        color = "black", linewidth = 1.4, inherit.aes = FALSE
      )
  }

  p
}

samples <- sort(unique(Combined_Positions_Relative$SampleID))

for (s in samples) {
  ggsave(
    filename = paste0(s, "_Rel_MDAMB231_withHulls_fixed300.png"),
    plot     = plot_sample_with_hulls(s, "MDAMB231"),
    width    = 7, height = 6, dpi = 600
  )
}
```

#### Grouped on one plot
```{r}
hull_df <- function(df, x = "RelX", y = "RelY") {
  pts <- df[, c(x, y), drop = FALSE]
  pts[[x]] <- suppressWarnings(as.numeric(pts[[x]]))
  pts[[y]] <- suppressWarnings(as.numeric(pts[[y]]))
  pts <- pts[is.finite(pts[[x]]) & is.finite(pts[[y]]), , drop = FALSE]
  pts <- dplyr::distinct(pts)

  if (nrow(pts) < 3) return(NULL)

  idx <- chull(pts[[x]], pts[[y]])
  out <- pts[idx, , drop = FALSE]
  out <- rbind(out, out[1, , drop = FALSE])  # close polygon
  names(out) <- c("RelX", "RelY")
  out
}

build_sample_hulls <- function(dat, cell_filter = "MDAMB231") {
  df <- dat %>%
    dplyr::filter(CellTypeID != "Missing_ID") %>%
    dplyr::mutate(
      RelX = suppressWarnings(as.numeric(RelX)),
      RelY = suppressWarnings(as.numeric(RelY))
    ) %>%
    dplyr::filter(is.finite(RelX), is.finite(RelY))

  if (!identical(cell_filter, "ALL")) {
    df <- df %>% dplyr::filter(CellTypeID == cell_filter)
  }

  df %>%
    dplyr::group_by(Group, SampleID) %>%
    dplyr::summarise(
      poly = list(hull_df(dplyr::pick(dplyr::everything()), "RelX", "RelY")),
      .groups = "drop"
    ) %>%
    dplyr::filter(!purrr::map_lgl(poly, is.null)) %>%
    tidyr::unnest(poly)
}

plot_group_hull_overlays <- function(hulls_df,
                                     group_name = c("Segregated", "Clustered"),
                                     fixed_limits = c(-300, 300, -300, 300),
                                     pad = 12,
                                     title = NULL,
                                     show_legend = TRUE) {

  group_name <- match.arg(group_name)

  h <- hulls_df %>% dplyr::filter(Group == group_name)
  if (!nrow(h)) return(ggplot() + ggtitle(paste0(group_name, " — EMPTY")))

  samples <- sort(unique(h$SampleID))
  repnum  <- stringr::str_extract(samples, "\\d+$")
  style_by_rep <- c("1" = "solid", "2" = "longdash", "3" = "1212")
  lt_vals <- style_by_rep[repnum]
  lt_vals[is.na(lt_vals)] <- "solid"
  lt_map <- setNames(lt_vals, samples)

  outline_col <- if (!is.null(celltype_colors[["MDAMB231"]])) celltype_colors[["MDAMB231"]] else "#B64249"
  fill_cols_h <- setNames(rep(outline_col, length(samples)), samples)

  xlim <- c(fixed_limits[1] - pad, fixed_limits[2] + pad)
  ylim <- c(fixed_limits[3] - pad, fixed_limits[4] + pad)

  ggplot(h, aes(x = RelX, y = RelY, group = SampleID)) +
    geom_polygon(aes(fill = SampleID), alpha = 0.22, color = NA) +
    geom_path(aes(linetype = SampleID), color = outline_col, linewidth = 1.6) +
    coord_fixed(xlim = xlim, ylim = ylim, expand = FALSE, clip = "off") +
    scale_linetype_manual(values = lt_map, drop = FALSE) +
    scale_fill_manual(values = fill_cols_h, drop = FALSE) +
    scale_x_continuous(breaks = c(-300, -150, 0, 150, 300), expand = c(0, 0)) +
    scale_y_continuous(breaks = c(-300, -150, 0, 150, 300), expand = c(0, 0)) +
    labs(
      title    = if (is.null(title)) paste0("MDAMB231 — ", group_name, " Sample Hull Overlays") else title,
      x        = "Relative X Position (µm)",
      y        = "Relative Y Position (µm)",
      linetype = "Sample",
      fill     = "Sample"
    ) +
    theme_classic(base_size = 22) +
    theme(
      legend.position = if (show_legend) "right" else "none",
      axis.title = element_text(size = 20),
      axis.text  = element_text(size = 18),
      plot.title = element_text(size = 22, face = "bold"),
      plot.margin = grid::unit(c(12, 18, 12, 12), "pt")
    )
}

sample_hulls_MDA <- build_sample_hulls(Combined_Positions_Relative, cell_filter = "MDAMB231")

p_seg_hull <- plot_group_hull_overlays(
  sample_hulls_MDA, "Segregated",
  fixed_limits = c(-300, 300, -300, 300),
  pad = 12,
  title = "MDAMB231 — Segregated (Sample Hull Overlays)",
  show_legend = TRUE
)

p_clus_hull <- plot_group_hull_overlays(
  sample_hulls_MDA, "Clustered",
  fixed_limits = c(-300, 300, -300, 300),
  pad = 12,
  title = "MDAMB231 — Clustered (Sample Hull Overlays)",
  show_legend = TRUE
)

ggsave("ConvexHulls_MDAMB231_Segregated_Overlays.png", p_seg_hull,  width = 7, height = 6, dpi = 600)
ggsave("ConvexHulls_MDAMB231_Clustered_Overlays.png",  p_clus_hull, width = 7, height = 6, dpi = 600)

p_seg_hull
p_clus_hull
```

#### Stats
```{r}
SampleHullAreas <- Combined_Positions_Relative %>%
  dplyr::filter(CellTypeID == "MDAMB231", Group %in% c("Segregated", "Clustered")) %>%
  dplyr::mutate(
    RelX = suppressWarnings(as.numeric(RelX)),
    RelY = suppressWarnings(as.numeric(RelY))
  ) %>%
  dplyr::filter(is.finite(RelX), is.finite(RelY)) %>%
  dplyr::group_by(SampleID, Group) %>%
  dplyr::summarise(
    hull_area = {
      pts <- cbind(RelX, RelY)
      pts <- pts[complete.cases(pts), , drop = FALSE]
      if (nrow(pts) < 3) NA_real_ else geometry::convhulln(pts, "FA")$area
    },
    .groups = "drop"
  ) %>%
  dplyr::mutate(Group = factor(Group, levels = c("Segregated", "Clustered")))

tt_dat <- SampleHullAreas %>%
  dplyr::filter(!is.na(hull_area), is.finite(hull_area), hull_area > 0)

t.test(hull_area ~ Group, data = tt_dat, var.equal = FALSE)
```

#### Per sample quant
```{r}
df_plot <- SampleHullAreas %>%
  dplyr::filter(!is.na(hull_area), is.finite(hull_area), hull_area > 0) %>%
  dplyr::mutate(Group = factor(Group, levels = c("Segregated", "Clustered")))

star_from_p <- function(p) {
  if (is.na(p)) return(NA_character_)
  if (p < 0.0001) return("****")
  if (p < 0.001)  return("***")
  if (p < 0.01)   return("**")
  if (p < 0.05)   return("*")
  "ns"
}

tt   <- t.test(hull_area ~ Group, data = df_plot, var.equal = FALSE)
pval <- tt$p.value
sig  <- star_from_p(pval)

data_max <- max(df_plot$hull_area, na.rm = TRUE)

# bracket position
base_y <- data_max * 1.06

stat_df_hull <- tibble::tibble(
  group1     = "Segregated",
  group2     = "Clustered",
  y.position = base_y,
  label      = sig
) %>%
  dplyr::filter(!is.na(label), label != "ns")

ylim_top_hull <- if (nrow(stat_df_hull) > 0) {
  max(stat_df_hull$y.position, na.rm = TRUE) * 1.08
} else {
  data_max * 1.05
}

# breaks: 250 if under ~2000 else 500
step <- if (ylim_top_hull <= 2000) 250 else 500
ymax_break <- ceiling(ylim_top_hull / step) * step
y_breaks <- seq(0, ymax_break, by = step)

p_box <- ggplot(df_plot, aes(x = Group, y = hull_area)) +
  geom_boxplot(
    aes(fill = Group),
    width         = 0.55,
    outlier.shape = NA,
    alpha         = 0.85,
    linewidth     = 0.8
  ) +
  geom_jitter(
    aes(color = Group),
    width = 0.15,
    size  = 3,
    alpha = 0.8,
    show.legend = FALSE
  ) +
  scale_x_discrete(labels = c("Segregated" = "Segregated", "Clustered" = "Clustered")) +
  scale_fill_manual(values = group_colors, drop = FALSE) +
  scale_color_manual(values = point_colors, drop = FALSE) +
  scale_y_continuous(breaks = y_breaks, expand = c(0, 0)) +
  coord_cartesian(ylim = c(0, ylim_top_hull)) +
  {
    if (nrow(stat_df_hull) > 0) {
      ggpubr::stat_pvalue_manual(
        stat_df_hull,
        xmin = "group1",
        xmax = "group2",
        label = "label",
        y.position = "y.position",
        tip.length   = 0.03,
        size         = 14,
        bracket.size = 0.9
      )
    } else {
      NULL
    }
  } +
  labs(
    title = "MDAMB231",
    x     = "Spatial Arrangement",
    y     = expression("TNBC Migration Territory ("*mu*m^2*")")
  ) +
  theme_classic(base_size = 22) +
  theme(
    axis.line    = element_line(linewidth = 0.9, color = "black"),
    axis.text.y  = element_text(size = 18),
    axis.title.y = element_text(size = 20),
    axis.text.x  = element_text(size = 20, angle = 0, hjust = 0.5),
    axis.title.x = element_text(size = 18),
    plot.title   = element_text(size = 16, hjust = 0.5),
    legend.position = "none"
  )

ggsave(
  "HullArea_MDAMB231_Box_Seg_vs_Clus.png",
  p_box,
  width  = 6,
  height = 6,
  dpi    = 600
)

p_box
```


### 3.) Track Length
```{r}
## 1) Numeric coords + sort (match your parsing style)
CP_num_len <- Combined_Positions %>%
  dplyr::mutate(
    TimeNum = suppressWarnings(readr::parse_number(Time)),
    X       = suppressWarnings(readr::parse_number(X_Position)),
    Y       = suppressWarnings(readr::parse_number(Y_Position))
  ) %>%
  dplyr::filter(is.finite(TimeNum), is.finite(X), is.finite(Y)) %>%
  dplyr::arrange(SampleID, UniqueID, TimeNum)

## 2) Track length per track (UniqueID)
TrackLengths <- CP_num_len %>%
  dplyr::group_by(SampleID, UniqueID, Group, CellTypeID) %>%
  dplyr::summarise(
    TrackLength = {
      x <- X
      y <- Y
      ok <- complete.cases(x, y)
      x <- x[ok]; y <- y[ok]
      if (length(x) < 2) {
        NA_real_
      } else {
        sum(sqrt(diff(x)^2 + diff(y)^2))
      }
    },
    n_points = sum(complete.cases(X, Y)),
    .groups = "drop"
  )

## 3) Restrict to your 2 cell types + your 2 groups
df_len <- TrackLengths %>%
  dplyr::filter(
    CellTypeID %in% c("MDAMB231", "Myoepithelial"),
    Group %in% c("Segregated", "Clustered"),
    !is.na(TrackLength)
  ) %>%
  dplyr::mutate(
    Group      = factor(Group, levels = c("Segregated", "Clustered")),
    CellTypeID = factor(CellTypeID, levels = c("MDAMB231", "Myoepithelial"))
  )
```

#### Stats
```{r}
welch_tracklength_by_celltype <- function(df) {

  star_from_p <- function(p) {
    if (is.na(p)) return(NA_character_)
    if (p <= 0.0001) return("****")
    if (p <= 0.001)  return("***")
    if (p <= 0.01)   return("**")
    if (p <= 0.05)   return("*")
    "ns"
  }

  df %>%
    dplyr::group_by(CellTypeID) %>%
    dplyr::group_modify(function(d, key) {

      d <- d %>%
        tidyr::drop_na(TrackLength, Group) %>%
        dplyr::mutate(Group = factor(Group, levels = c("Segregated", "Clustered")))

      # require both groups + >=2 tracks per group
      if (dplyr::n_distinct(d$Group) < 2 || any(table(d$Group) < 2)) {
        return(tibble::tibble(
          comparison = "Segregated-Clustered",
          direction  = NA_character_,
          p_value    = NA_real_,
          sig        = NA_character_,
          statement  = paste0(as.character(key$CellTypeID), ": Not enough tracks in one group")
        ))
      }

      tt <- t.test(TrackLength ~ Group, data = d, var.equal = FALSE)
      p  <- tt$p.value

      means <- tapply(d$TrackLength, d$Group, mean, na.rm = TRUE)
      dir_txt <- if (means["Clustered"] > means["Segregated"]) "Clustered > Segregated" else "Segregated > Clustered"

      tibble::tibble(
        comparison = "Segregated-Clustered",
        direction  = dir_txt,
        p_value    = p,
        sig        = star_from_p(p),
        statement  = paste0(
          as.character(key$CellTypeID),
          " TrackLength: ", dir_txt,
          " (Welch t-test, p = ", signif(p, 3), ", ", star_from_p(p), ")"
        )
      )
    }) %>%
    dplyr::ungroup()
}

tracklen_pairwise <- welch_tracklength_by_celltype(df_len)

tracklen_pairwise
tracklen_pairwise$statement
```


#### Viz
```{r}
fill_cols  <- group_colors
point_cols <- point_colors

data_max <- max(df_len$TrackLength, na.rm = TRUE)
step <- if (data_max <= 2000) 250 else 500

sig_rows <- tracklen_pairwise %>%
  dplyr::filter(!is.na(sig), sig != "ns") %>%
  dplyr::mutate(
    group1 = "Segregated",
    group2 = "Clustered"
  )

ymax_ceil <- ceiling(data_max / step) * step
base_y <- ymax_ceil * 0.80
step_y <- ymax_ceil * 0.08

stat_df_len <- sig_rows %>%
  dplyr::group_by(CellTypeID) %>%
  dplyr::mutate(
    y.position = base_y + step_y * (dplyr::row_number() - 1),
    label      = sig
  ) %>%
  dplyr::ungroup() %>%
  dplyr::select(CellTypeID, group1, group2, y.position, label)

bracket_max <- if (nrow(stat_df_len) > 0) max(stat_df_len$y.position, na.rm = TRUE) else 0

# probe violin ymax so brackets never get clipped
p_probe_len <- ggplot(df_len, aes(x = Group, y = TrackLength)) +
  geom_violin(trim = FALSE, width = 0.95)

violin_max_len <- max(ggplot_build(p_probe_len)$data[[1]]$ymax, na.rm = TRUE)
ylim_top_len <- max(violin_max_len, bracket_max) * 1.05

ymax_break <- ceiling(ylim_top_len / step) * step
y_breaks   <- seq(0, ymax_break, by = step)

p_len <- ggplot(df_len, aes(x = Group, y = TrackLength)) +
  geom_violin(
    aes(fill = Group),
    trim  = FALSE,
    alpha = 0.35,
    color = NA,
    width = 0.95
  ) +
  geom_boxplot(
    aes(fill = Group),
    width         = 0.22,
    outlier.shape = NA,
    alpha         = 0.85,
    linewidth     = 0.8
  ) +
  geom_jitter(
    aes(color = Group),
    width = 0.15,
    size  = 2,
    alpha = 0.8,
    show.legend = FALSE
  ) +
  facet_grid(. ~ CellTypeID) +
  scale_fill_manual(values = fill_cols, drop = FALSE) +
  scale_color_manual(values = point_cols, drop = FALSE) +
  scale_y_continuous(breaks = y_breaks, expand = c(0, 0)) +
  scale_x_discrete(labels = c("Segregated" = "Segregated", "Clustered" = "Clustered")) +
  coord_cartesian(ylim = c(0, ylim_top_len)) +
  {
    if (nrow(stat_df_len) > 0) {
      ggpubr::stat_pvalue_manual(
        stat_df_len,
        xmin = "group1",
        xmax = "group2",
        label = "label",
        y.position = "y.position",
        tip.length   = 0.03,
        size         = 6,
        bracket.size = 0.9
      )
    } else {
      NULL
    }
  } +
  labs(
    title = NULL,
    x     = "Spatial Arrangement",
    y     = "Total Distance Traveled (µm)"
  ) +
  theme_classic(base_size = 20) +
  theme(
    axis.line    = element_line(linewidth = 0.9, color = "black"),
    axis.text.y  = element_text(size = 18),
    axis.title.y = element_text(size = 20),
    axis.text.x  = element_text(size = 15, hjust = 0.5),
    axis.title.x = element_text(size = 14),
    legend.position = "none",
    strip.background = element_blank(),
    strip.text.x = element_text(size = 20),
    panel.spacing.x = unit(1.5, "lines")
  )

ggsave(
  filename = "TrackLength_Seg_Clus.png",
  plot     = p_len,
  width    = 6.5,
  height   = 4.7,
  dpi      = 600
)

p_len
```


### 4.) Normalized Distance (Distance ÷ Frames) Accounts for tracks truncated from cell divisions and cells that go out of frame.
```{r}
CP_num_dist <- Combined_Positions %>%
  mutate(
    TimeNum = suppressWarnings(readr::parse_number(Time)),
    X       = suppressWarnings(readr::parse_number(X_Position)),
    Y       = suppressWarnings(readr::parse_number(Y_Position))
  ) %>%
  filter(is.finite(TimeNum), is.finite(X), is.finite(Y)) %>%
  arrange(SampleID, UniqueID, TimeNum)

# 2) Distance per frame (normalizes for divisions / out-of-frame)
DistPerFrame_df <- CP_num_dist %>%
  filter(
    CellTypeID %in% c("MDAMB231", "Myoepithelial"),
    Group %in% c("Segregated", "Clustered")
  ) %>%
  group_by(SampleID, UniqueID, Group, CellTypeID) %>%
  summarise(
    n_frames_visible = n(),

    TrackLength = {
      x <- X
      y <- Y
      if (length(x) < 2) NA_real_ else sum(sqrt(diff(x)^2 + diff(y)^2))
    },
    .groups = "drop"
  ) %>%
  filter(!is.na(TrackLength), n_frames_visible >= 2) %>%
  mutate(
    Group = factor(Group, levels = c("Segregated", "Clustered")),
    CellTypeID = factor(CellTypeID, levels = c("MDAMB231", "Myoepithelial")),
    Dist_um_per_frame = TrackLength / (n_frames_visible - 1)
  )

df_dist_frame <- DistPerFrame_df

# sanity
summary(df_dist_frame$Dist_um_per_frame)
```

#### Stats
```{r}
welch_distframe_by_celltype <- function(df) {

  star_from_p <- function(p) {
    if (is.na(p)) return(NA_character_)
    if (p <= 0.0001) return("****")
    if (p <= 0.001)  return("***")
    if (p <= 0.01)   return("**")
    if (p <= 0.05)   return("*")
    "ns"
  }

  df %>%
    group_by(CellTypeID) %>%
    group_modify(function(d, key) {

      d <- d %>%
        tidyr::drop_na(Dist_um_per_frame, Group) %>%
        mutate(Group = factor(Group, levels = c("Segregated", "Clustered")))

      # require both groups + >=2 tracks per group
      if (n_distinct(d$Group) < 2 || any(table(d$Group) < 2)) {
        return(tibble(
          comparison = "Segregated-Clustered",
          direction  = NA_character_,
          p_value    = NA_real_,
          sig        = NA_character_,
          statement  = paste0(as.character(key$CellTypeID), ": Not enough tracks in one group")
        ))
      }

      tt <- t.test(Dist_um_per_frame ~ Group, data = d, var.equal = FALSE)
      p  <- tt$p.value

      means <- tapply(d$Dist_um_per_frame, d$Group, mean, na.rm = TRUE)
      dir_txt <- if (means["Clustered"] > means["Segregated"]) {
        "Clustered > Segregated"
      } else {
        "Segregated > Clustered"
      }

      tibble(
        comparison = "Segregated-Clustered",
        direction  = dir_txt,
        p_value    = p,
        sig        = star_from_p(p),
        statement  = paste0(
          as.character(key$CellTypeID),
          " Distance/frame: ", dir_txt,
          " (Welch t-test, p = ", signif(p, 3), ", ", star_from_p(p), ")"
        )
      )
    }) %>%
    ungroup()
}

dist_frame_pairwise <- welch_distframe_by_celltype(df_dist_frame)

dist_frame_pairwise
dist_frame_pairwise$statement
```

#### Viz
```{r}
fill_cols  <- group_colors
point_cols <- point_colors

if (nrow(df_dist_frame) == 0) {
  stop("df_dist_frame is empty. Check Group / CellTypeID filtering upstream.")
}

sig_rows <- dist_frame_pairwise %>%
  filter(!is.na(sig), sig != "ns") %>%
  mutate(group1 = "Segregated", group2 = "Clustered")

# --- FIXED AXIS RANGE ---
ylim_top  <- 18
y_breaks  <- seq(0, 18, by = 2.5)

# bracket height (kept relative but capped inside axis)
stat_df_dist <- sig_rows %>%
  group_by(CellTypeID) %>%
  mutate(
    data_max   = max(df_dist_frame$Dist_um_per_frame[df_dist_frame$CellTypeID == CellTypeID[1]], na.rm = TRUE),
    y.position = pmin(data_max * 1.10, 17.2),
    label      = sig
  ) %>%
  ungroup() %>%
  select(CellTypeID, group1, group2, y.position, label)

p_dist_frame <- ggplot(df_dist_frame, aes(x = Group, y = Dist_um_per_frame)) +
  geom_violin(aes(fill = Group), trim = FALSE, alpha = 0.35, color = NA, width = 0.95) +
  geom_boxplot(aes(fill = Group), width = 0.22, outlier.shape = NA, alpha = 0.85, linewidth = 0.8) +
  geom_jitter(aes(color = Group), width = 0.15, size = 2, alpha = 0.8, show.legend = FALSE) +
  facet_grid(. ~ CellTypeID) +
  scale_fill_manual(values = fill_cols, drop = FALSE) +
  scale_color_manual(values = point_cols, drop = FALSE) +
  scale_x_discrete(labels = c("Segregated" = "Segregated", "Clustered" = "Clustered")) +
  scale_y_continuous(breaks = y_breaks, limits = c(0, 18), expand = c(0, 0)) +
  coord_cartesian(clip = "off") +
  {
    if (nrow(stat_df_dist) > 0) {
      ggpubr::stat_pvalue_manual(
        stat_df_dist,
        xmin = "group1",
        xmax = "group2",
        label = "label",
        y.position = "y.position",
        tip.length = 0.03,
        size = 14,
        bracket.size = 0.9
      )
    } else NULL
  } +
  labs(
    title = NULL,
    x     = "Spatial Arrangement",
    y     = "Normalized Distance \nTraveled (µm / Frame)"
  ) +
  theme_classic(base_size = 20) +
  theme(
    axis.line    = element_line(linewidth = 0.9, color = "black"),
    axis.text.y  = element_text(size = 18),
    axis.title.y = element_text(size = 20),
    axis.text.x  = element_text(size = 16, hjust = 0.5),
    axis.title.x = element_text(size = 16),
    legend.position = "none",
    strip.background = element_blank(),
    strip.text.x = element_text(size = 20),
    panel.spacing.x = unit(2, "lines"),
    plot.margin = margin(10, 10, 10, 10)
  )

ggsave(
  "DistancePerVisibleFrame_Seg_Clus.png",
  p_dist_frame,
  width  = 7,
  height = 5,
  dpi    = 600
)

p_dist_frame
```

### 5.) Net Displacement
```{r}
fill_cols  <- group_colors
point_cols <- point_colors

disp_df <- Combined_Positions %>%
  mutate(
    TimeNum = suppressWarnings(readr::parse_number(as.character(Time))),
    X       = suppressWarnings(readr::parse_number(as.character(X_Position))),
    Y       = suppressWarnings(readr::parse_number(as.character(Y_Position)))
  ) %>%
  filter(
    is.finite(TimeNum), is.finite(X), is.finite(Y),
    Group %in% c("Segregated", "Clustered"),
    CellTypeID %in% c("MDAMB231", "Myoepithelial"),
    CellTypeID != "Missing_ID"
  ) %>%
  arrange(SampleID, UniqueID, TimeNum) %>%
  group_by(SampleID, UniqueID) %>%
  summarise(
    Group      = first(Group),
    CellTypeID = first(CellTypeID),

    # sanity checks (drop broken tracks)
    group_n    = n_distinct(Group),
    celltype_n = n_distinct(CellTypeID),

    x0 = first(X),
    y0 = first(Y),
    x1 = last(X),
    y1 = last(Y),

    Displacement_um = sqrt((x1 - x0)^2 + (y1 - y0)^2),
    n_points = n(),
    .groups = "drop"
  ) %>%
  filter(group_n == 1, celltype_n == 1, n_points >= 2) %>%
  select(-group_n, -celltype_n) %>%
  mutate(
    Group      = factor(Group, levels = c("Segregated", "Clustered")),
    CellTypeID = factor(CellTypeID, levels = c("MDAMB231", "Myoepithelial"))
  )

# optional: quick counts
disp_df %>% count(CellTypeID, Group)
```

#### Stats
```{r}
welch_displacement_by_celltype <- function(df) {

  star_from_p <- function(p) {
    if (is.na(p)) return(NA_character_)
    if (p <= 0.0001) return("****")
    if (p <= 0.001)  return("***")
    if (p <= 0.01)   return("**")
    if (p <= 0.05)   return("*")
    "ns"
  }

  df %>%
    group_by(CellTypeID) %>%
    group_modify(function(d, key) {

      d <- d %>%
        tidyr::drop_na(Displacement_um, Group) %>%
        mutate(Group = factor(Group, levels = c("Segregated", "Clustered")))

      # require both groups + >=2 tracks per group
      if (n_distinct(d$Group) < 2 || any(table(d$Group) < 2)) {
        return(tibble(
          comparison = "Segregated-Clustered",
          direction  = NA_character_,
          p_value    = NA_real_,
          sig        = NA_character_,
          statement  = paste0(as.character(key$CellTypeID), ": Not enough tracks in one group")
        ))
      }

      tt <- t.test(Displacement_um ~ Group, data = d, var.equal = FALSE)
      p  <- tt$p.value

      means <- tapply(d$Displacement_um, d$Group, mean, na.rm = TRUE)
      dir_txt <- if (means["Clustered"] > means["Segregated"]) {
        "Clustered > Segregated"
      } else {
        "Segregated > Clustered"
      }

      tibble(
        comparison = "Segregated-Clustered",
        direction  = dir_txt,
        p_value    = p,
        sig        = star_from_p(p),
        statement  = paste0(
          as.character(key$CellTypeID),
          " Net Displacement: ", dir_txt,
          " (Welch t-test, p = ", signif(p, 3), ", ", star_from_p(p), ")"
        )
      )
    }) %>%
    ungroup()
}

disp_pairwise <- welch_displacement_by_celltype(disp_df)

disp_pairwise
disp_pairwise$statement
```

#### Viz
```{r}
sig_rows <- disp_pairwise %>%
  filter(!is.na(sig), sig != "ns") %>%
  mutate(group1 = "Segregated", group2 = "Clustered")

data_max <- max(disp_df$Displacement_um, na.rm = TRUE)
step <- if (data_max <= 500) 50 else if (data_max <= 1000) 100 else if (data_max <= 2000) 250 else 500

ymax_ceil <- ceiling(data_max / step) * step

# bracket heights per facet (not hard-coded; stays above the data)
stat_df_disp <- sig_rows %>%
  group_by(CellTypeID) %>%
  mutate(
    data_max_ct = max(disp_df$Displacement_um[disp_df$CellTypeID == CellTypeID[1]], na.rm = TRUE),
    y.position  = data_max_ct * 1.3,
    label       = sig
  ) %>%
  ungroup() %>%
  select(CellTypeID, group1, group2, y.position, label)

ylim_top <- if (nrow(stat_df_disp) > 0) {
  max(stat_df_disp$y.position, na.rm = TRUE) * 1.05
} else {
  ymax_ceil * 1.10
}

ymax_break <- ceiling(ylim_top / step) * step
y_breaks   <- seq(0, ymax_break, by = step)

p_disp <- ggplot(disp_df, aes(x = Group, y = Displacement_um)) +
  geom_violin(aes(fill = Group), trim = FALSE, alpha = 0.35, color = NA, width = 0.95) +
  geom_boxplot(aes(fill = Group), width = 0.22, outlier.shape = NA, alpha = 0.85, linewidth = 0.8) +
  geom_jitter(aes(color = Group), width = 0.15, size = 2, alpha = 0.8, show.legend = FALSE) +
  facet_grid(. ~ CellTypeID) +
  scale_fill_manual(values = fill_cols, drop = FALSE) +
  scale_color_manual(values = point_cols, drop = FALSE) +
  scale_x_discrete(labels = c("Segregated" = "Segregated", "Clustered" = "Clustered")) +
  scale_y_continuous(breaks = y_breaks, expand = expansion(mult = c(0, 0.02))) +
  coord_cartesian(ylim = c(0, ylim_top), clip = "off") +
  {
    if (nrow(stat_df_disp) > 0) {
      ggpubr::stat_pvalue_manual(
        stat_df_disp,
        xmin = "group1",
        xmax = "group2",
        label = "label",
        y.position = "y.position",
        tip.length = 0.03,
        size = 12,
        bracket.size = 0.9
      )
    } else NULL
  } +
  labs(
    title = NULL,
    x = "Spatial Arrangement",
    y = "Net Displacement (µm)"
  ) +
  theme_classic(base_size = 20) +
  theme(
    axis.line    = element_line(linewidth = 0.9, color = "black"),
    axis.text.y  = element_text(size = 18),
    axis.title.y = element_text(size = 20),
    axis.text.x  = element_text(size = 15, angle = 0, hjust = 0.5),
    axis.title.x = element_text(size = 14),
    legend.position = "none",
    strip.background = element_blank(),
    strip.text.x = element_text(size = 20),
    panel.spacing.x = unit(1.5, "lines"),
    plot.margin = margin(10, 10, 10, 10)
  )

ggsave(
  filename = "NetDisplacement_Seg_Clus.png",
  plot     = p_disp,
  width    = 6.5,
  height   = 4.7,
  dpi      = 600
)

p_disp
```

### 6.) MSD
```{r}
frame_interval_min <- 10

prep_tracks_msd <- function(df) {
  df %>%
    dplyr::mutate(
      TimeNum = suppressWarnings(readr::parse_number(as.character(Time))),
      X       = suppressWarnings(readr::parse_number(as.character(X_Position))),
      Y       = suppressWarnings(readr::parse_number(as.character(Y_Position)))
    ) %>%
    dplyr::filter(is.finite(TimeNum), is.finite(X), is.finite(Y)) %>%
    dplyr::arrange(UniqueID, TimeNum) %>%
    dplyr::distinct(UniqueID, TimeNum, .keep_all = TRUE)
}

msd_per_track <- function(track_df) {
  x <- track_df$X
  y <- track_df$Y
  n <- length(x)
  if (n < 2) return(tibble::tibble(tau_frames = integer(), msd_track = numeric()))

  k_vec <- seq_len(n - 1)
  msd_k <- vapply(k_vec, function(k) {
    dx <- x[(1 + k):n] - x[1:(n - k)]
    dy <- y[(1 + k):n] - y[1:(n - k)]
    mean(dx*dx + dy*dy, na.rm = TRUE)
  }, numeric(1))

  tibble::tibble(tau_frames = k_vec, msd_track = msd_k)
}

calc_msd_unbalanced <- function(df, frame_interval_min = 10, min_tracks_for_sem = 3) {
  df <- df %>%
    prep_tracks_msd() %>%
    dplyr::group_by(UniqueID) %>%
    dplyr::arrange(TimeNum, .by_group = TRUE)

  if (nrow(df) == 0) {
    return(tibble::tibble(
      tau_frames = integer(),
      tau_min    = numeric(),
      tau_hr     = numeric(),
      msd        = numeric(),
      sem        = numeric(),
      n_tracks   = integer()
    ))
  }

  per_track <- df %>%
    dplyr::group_split(UniqueID, .keep = FALSE) %>%
    purrr::map_dfr(msd_per_track, .id = "track_index")

  per_track %>%
    dplyr::group_by(tau_frames) %>%
    dplyr::summarise(
      msd      = mean(msd_track, na.rm = TRUE),
      sem      = sd(msd_track, na.rm = TRUE) / sqrt(sum(is.finite(msd_track))),
      n_tracks = sum(is.finite(msd_track)),
      .groups  = "drop"
    ) %>%
    dplyr::mutate(
      tau_min = tau_frames * frame_interval_min,
      tau_hr  = tau_min / 60,
      sem     = ifelse(n_tracks >= min_tracks_for_sem, sem, NA_real_)
    ) %>%
    dplyr::arrange(tau_frames)
}

compute_msd_by_celltype <- function(dat,
                                    celltype,
                                    groups = c("Segregated", "Clustered"),
                                    frame_interval_min = 10,
                                    min_tracks_for_sem = 3) {
  purrr::map_dfr(groups, function(g) {
    subset_df <- dat %>%
      dplyr::filter(CellTypeID == celltype, Group == g, CellTypeID != "Missing_ID") %>%
      dplyr::select(UniqueID, Time, X_Position, Y_Position)

    out <- calc_msd_unbalanced(
      subset_df,
      frame_interval_min = frame_interval_min,
      min_tracks_for_sem = min_tracks_for_sem
    )
    out$Group      <- g
    out$CellTypeID <- celltype
    out
  })
}

trim_to_common_tau <- function(summary_df, groups = c("Segregated", "Clustered")) {
  df <- summary_df %>% dplyr::filter(Group %in% groups)

  max_tau_by_group <- df %>%
    dplyr::group_by(CellTypeID, Group) %>%
    dplyr::summarise(
      max_tau = max(tau_frames[n_tracks >= 1], na.rm = TRUE),
      .groups = "drop"
    )

  common_tau_by_ct <- max_tau_by_group %>%
    dplyr::group_by(CellTypeID) %>%
    dplyr::summarise(common_tau = min(max_tau, na.rm = TRUE), .groups = "drop")

  df %>%
    dplyr::left_join(common_tau_by_ct, by = "CellTypeID") %>%
    dplyr::filter(is.finite(common_tau), tau_frames <= common_tau) %>%
    dplyr::select(-common_tau)
}

# --- Compute MSD summaries (Segregated first, Clustered second) ---
msd_mda <- compute_msd_by_celltype(
  Combined_Positions,
  celltype = "MDAMB231",
  groups = c("Segregated", "Clustered"),
  frame_interval_min = frame_interval_min,
  min_tracks_for_sem = 3
)

msd_myo <- compute_msd_by_celltype(
  Combined_Positions,
  celltype = "Myoepithelial",
  groups = c("Segregated", "Clustered"),
  frame_interval_min = frame_interval_min,
  min_tracks_for_sem = 3
)

msd_all <- dplyr::bind_rows(msd_mda, msd_myo) %>%
  dplyr::mutate(
    Group      = factor(Group, levels = c("Segregated", "Clustered")),
    CellTypeID = factor(CellTypeID, levels = c("MDAMB231", "Myoepithelial"))
  )

# Trim so Segregated and Clustered have identical x-range within each facet
msd_all_trim <- trim_to_common_tau(msd_all, groups = c("Segregated", "Clustered"))

# y-limit based on observed (msd + sem)
y_max <- max(msd_all_trim$msd + ifelse(is.finite(msd_all_trim$sem), msd_all_trim$sem, 0), na.rm = TRUE)
y_max <- ceiling(y_max / 1000) * 1000

# Ribbon only where SEM exists
msd_ribbon <- msd_all_trim %>% dplyr::filter(is.finite(sem))

# Nice x breaks in hours (every 6 h, up to max)
xmax_hr <- max(msd_all_trim$tau_hr, na.rm = TRUE)
x_breaks_hr <- seq(0, ceiling(xmax_hr / 6) * 6, by = 6)

# --- Plot ---
p_msd <- ggplot(msd_all_trim, aes(x = tau_hr, y = msd, color = Group, fill = Group)) +
  geom_ribbon(
    data = msd_ribbon,
    aes(ymin = pmax(msd - sem, 0), ymax = msd + sem),
    alpha = 0.18,
    color = NA
  ) +
  geom_line(linewidth = 0.9, na.rm = TRUE) +
  geom_point(size = 0.5, alpha = 0.6, na.rm = TRUE) +
  facet_grid(. ~ CellTypeID) +
  scale_color_manual(
    values = group_colors,
    breaks = c("Segregated", "Clustered"),
    labels = c("Segregated", "Clustered"),
    name   = "Spatial\nArrangement",
    drop   = FALSE
  ) +
  scale_fill_manual(
    values = group_colors,
    breaks = c("Segregated", "Clustered"),
    labels = c("Segregated", "Clustered"),
    name   = "Spatial\nArrangement",
    drop   = FALSE
  ) +
  scale_x_continuous(
    limits = c(0, NA),
    breaks = x_breaks_hr,
    expand = c(0, 0)
  ) +
  scale_y_continuous(limits = c(0, 200000), expand = expansion(mult = c(0, 0.02))) +
  labs(
    title = NULL,
    x = "Lag τ (hours)",
    y = expression("MSD ("*mu*"m"^2*")")
  ) +
  theme_classic(base_size = 22) +
  theme(
    legend.position  = "right",
    legend.title = element_text(size = 18, margin = margin(b = 4)),
    legend.text = element_text(size = 16),
    legend.key.height = unit(10, "pt"),
    legend.spacing.y  = unit(2, "pt"),
    strip.background = element_blank(),
    strip.text.x     = element_text(size = 24),
    axis.line        = element_line(linewidth = 1.2, color = "black"),
    axis.text        = element_text(size = 22),
    axis.title       = element_text(size = 24),
    panel.spacing.x  = grid::unit(2, "cm")
  )

ggsave(
  filename = "MSD_Segregated_Clustered_MDAMB231_Myo_sharedY.png",
  plot     = p_msd,
  width    = 12.5,
  height   = 6,
  dpi      = 600
)

p_msd
```

#### Alpha calculation
```{r}
fit_range_hr <- c(0.83, 8.33)   # 50 to 500 minutes expressed in hours

get_alpha <- function(df,
                      tau_col = "tau_hr",
                      msd_col = "msd",
                      tau_range = c(0.83, 8.33),
                      min_pts = 3) {

  stopifnot(tau_col %in% names(df), msd_col %in% names(df))

  df_filt <- df %>%
    dplyr::filter(
      is.finite(.data[[tau_col]]),
      .data[[tau_col]] >= tau_range[1],
      .data[[tau_col]] <= tau_range[2],
      is.finite(.data[[msd_col]]),
      .data[[msd_col]] > 0
    ) %>%
    dplyr::mutate(
      log_tau = log(.data[[tau_col]]),
      log_msd = log(.data[[msd_col]])
    )

  if (nrow(df_filt) == 0) {
    return(tibble::tibble(
      CellTypeID = character(),
      Group      = character(),
      n_pts      = integer(),
      alpha      = numeric(),
      r2         = numeric()
    ))
  }

  df_filt %>%
    dplyr::group_by(CellTypeID, Group) %>%
    tidyr::nest() %>%
    dplyr::mutate(
      n_pts  = purrr::map_int(data, nrow),
      model  = purrr::map(data, ~ if (nrow(.x) >= min_pts) stats::lm(log_msd ~ log_tau, data = .x) else NULL),
      alpha  = purrr::map_dbl(model, ~ if (is.null(.x)) NA_real_ else unname(stats::coef(.x)[2])),
      r2     = purrr::map_dbl(model, ~ if (is.null(.x)) NA_real_ else summary(.x)$r.squared)
    ) %>%
    dplyr::ungroup() %>%
    dplyr::mutate(
      Group = factor(Group, levels = c("Segregated", "Clustered"))
    ) %>%
    dplyr::arrange(CellTypeID, Group) %>%
    dplyr::select(CellTypeID, Group, n_pts, alpha, r2)
}

alpha_results <- get_alpha(
  msd_all_trim,
  tau_col   = "tau_hr",
  msd_col   = "msd",
  tau_range = fit_range_hr,
  min_pts   = 3
)

alpha_results %>%
  dplyr::mutate(
    alpha = sprintf("%.2f", alpha),
    r2    = sprintf("%.2f", r2)
  )

```


### 7.) Travel consistency
```{r}
# Travel consistency is net displacement / track length

# Build displacement df (we already have but if future code doesn't have here is how to do it )
if (!exists("disp_df")) {

  disp_df <- Combined_Positions %>%
    mutate(
      TimeNum = suppressWarnings(readr::parse_number(as.character(Time))),
      X       = suppressWarnings(readr::parse_number(as.character(X_Position))),
      Y       = suppressWarnings(readr::parse_number(as.character(Y_Position)))
    ) %>%
    filter(
      is.finite(TimeNum), is.finite(X), is.finite(Y),
      Group %in% c("Segregated", "Clustered"),
      CellTypeID %in% c("MDAMB231", "Myoepithelial"),
      CellTypeID != "Missing_ID"
    ) %>%
    arrange(SampleID, UniqueID, TimeNum) %>%
    group_by(SampleID, UniqueID) %>%
    summarise(
      Group      = first(Group),
      CellTypeID = first(CellTypeID),
      group_n    = n_distinct(Group),
      celltype_n = n_distinct(CellTypeID),
      x0 = first(X), y0 = first(Y),
      x1 = last(X),  y1 = last(Y),
      Displacement_um = sqrt((x1 - x0)^2 + (y1 - y0)^2),
      n_points = n(),
      .groups = "drop"
    ) %>%
    filter(group_n == 1, celltype_n == 1, n_points >= 2) %>%
    select(-group_n, -celltype_n, -n_points) %>%
    mutate(
      Group      = factor(Group, levels = c("Segregated", "Clustered")),
      CellTypeID = factor(CellTypeID, levels = c("MDAMB231", "Myoepithelial"))
    )
}

disp_df_use <- disp_df %>%
  filter(
    Group %in% c("Segregated", "Clustered"),
    CellTypeID %in% c("MDAMB231", "Myoepithelial"),
    is.finite(Displacement_um),
    Displacement_um >= 0
  ) %>%
  mutate(
    Group      = factor(Group, levels = c("Segregated", "Clustered")),
    CellTypeID = factor(CellTypeID, levels = c("MDAMB231", "Myoepithelial"))
  )

# Build track length df (we already have but if future code doesn't have here is how to do it )
if (!exists("TrackLengths")) {

  CP_num_len <- Combined_Positions %>%
    mutate(
      TimeNum = suppressWarnings(readr::parse_number(Time)),
      X       = suppressWarnings(readr::parse_number(X_Position)),
      Y       = suppressWarnings(readr::parse_number(Y_Position))
    ) %>%
    filter(is.finite(TimeNum), is.finite(X), is.finite(Y)) %>%
    arrange(SampleID, UniqueID, TimeNum)

  TrackLengths <- CP_num_len %>%
    group_by(SampleID, UniqueID, Group, CellTypeID) %>%
    summarise(
      TrackLength = {
        x <- X; y <- Y
        ok <- complete.cases(x, y)
        x <- x[ok]; y <- y[ok]
        if (length(x) < 2) NA_real_ else sum(sqrt(diff(x)^2 + diff(y)^2))
      },
      n_points = sum(complete.cases(X, Y)),
      .groups = "drop"
    )
}

TrackLengths_use <- TrackLengths %>%
  filter(
    Group %in% c("Segregated", "Clustered"),
    CellTypeID %in% c("MDAMB231", "Myoepithelial"),
    CellTypeID != "Missing_ID",
    is.finite(TrackLength),
    TrackLength > 0
  ) %>%
  mutate(
    Group      = factor(Group, levels = c("Segregated", "Clustered")),
    CellTypeID = factor(CellTypeID, levels = c("MDAMB231", "Myoepithelial"))
  )

# Merge + compute CI
CI_df <- TrackLengths_use %>%
  inner_join(
    disp_df_use %>% select(SampleID, UniqueID, Group, CellTypeID, Displacement_um),
    by = c("SampleID", "UniqueID", "Group", "CellTypeID")
  ) %>%
  mutate(
    CI = if_else(
      is.finite(TrackLength) & TrackLength > 0 &
        is.finite(Displacement_um) & Displacement_um >= 0,
      Displacement_um / TrackLength,
      NA_real_
    )
  ) %>%
  filter(is.finite(CI)) %>%
  mutate(
    CI = pmin(pmax(CI, 0), 1), # clamp tiny numerical overshoots
    Group      = factor(Group, levels = c("Segregated", "Clustered")),
    CellTypeID = factor(CellTypeID, levels = c("MDAMB231", "Myoepithelial"))
  )

# sanity
CI_df %>% count(CellTypeID, Group)
summary(CI_df$CI)
```

#### Stats
```{r}
pairwise_ci_by_celltype <- function(df) {

  df %>%
    group_by(CellTypeID) %>%
    group_modify(function(d, key) {

      d <- d %>%
        tidyr::drop_na(CI, Group) %>%
        mutate(Group = factor(Group, levels = c("Segregated", "Clustered")))

      # require both groups + >=2 tracks per group
      if (n_distinct(d$Group) < 2 || any(table(d$Group) < 2)) {
        return(tibble(
          comparison = "Segregated-Clustered",
          direction  = NA_character_,
          p_value    = NA_real_,
          sig        = NA_character_,
          statement  = paste0(as.character(key$CellTypeID), ": Not enough tracks in one group")
        ))
      }

      tt <- t.test(CI ~ Group, data = d, var.equal = FALSE)
      p  <- tt$p.value

      means <- tapply(d$CI, d$Group, mean, na.rm = TRUE)
      dir_txt <- if (means["Clustered"] > means["Segregated"]) {
        "Clustered > Segregated"
      } else {
        "Segregated > Clustered"
      }

      tibble(
        comparison = "Segregated-Clustered",
        direction  = dir_txt,
        p_value    = p,
        sig        = star_from_p(p),
        statement  = paste0(
          as.character(key$CellTypeID),
          " Travel Consistency: ", dir_txt,
          " (Welch t-test, p = ", signif(p, 3), ", ", star_from_p(p), ")"
        )
      )
    }) %>%
    ungroup()
}

ci_pairwise <- pairwise_ci_by_celltype(CI_df)

ci_pairwise
ci_pairwise$statement
```

#### Viz
```{r}
fill_cols  <- group_colors
point_cols <- point_colors

star_from_p <- function(p) {
  if (is.na(p)) return(NA_character_)
  if (p <= 0.0001) return("****")
  if (p <= 0.001)  return("***")
  if (p <= 0.01)   return("**")
  if (p <= 0.05)   return("*")
  "ns"
}

sig_rows <- ci_pairwise %>%
  filter(!is.na(sig), sig != "ns") %>%
  mutate(group1 = "Segregated", group2 = "Clustered")

# bracket heights per facet (data-driven)
stat_df_ci <- sig_rows %>%
  group_by(CellTypeID) %>%
  mutate(
    data_max_ct = max(CI_df$CI[CI_df$CellTypeID == CellTypeID[1]], na.rm = TRUE),

    # raise brackets a little for MDA, a lot for Myoep
    bump = dplyr::case_when(
      CellTypeID == "MDAMB231"      ~ 0.18,  
      CellTypeID == "Myoepithelial" ~ 0.3,  
      TRUE                          ~ 0.06
    ),

    y.position = pmin(1.35, data_max_ct + bump),
    label      = sig
  ) %>%
  ungroup() %>%
  select(CellTypeID, group1, group2, y.position, label)

p_ci <- ggplot(CI_df, aes(x = Group, y = CI)) +
  geom_violin(aes(fill = Group), trim = FALSE, alpha = 0.35, color = NA, width = 0.95) +
  geom_boxplot(aes(fill = Group), width = 0.22, outlier.shape = NA, alpha = 0.85, linewidth = 0.8) +
  geom_jitter(aes(color = Group), width = 0.15, size = 2, alpha = 0.55, show.legend = FALSE) +
  facet_grid(. ~ CellTypeID) +
  scale_fill_manual(values = fill_cols, drop = FALSE) +
  scale_color_manual(values = point_cols, drop = FALSE) +
  scale_x_discrete(labels = c("Segregated" = "Segregated", "Clustered" = "Clustered")) +
  scale_y_continuous(
    limits = c(0, 1.4),
    breaks = seq(0, 1.4, by = 0.2),
    expand = c(0, 0)
  ) +
  coord_cartesian(clip = "off") +
  {
    if (nrow(stat_df_ci) > 0) {
      ggpubr::stat_pvalue_manual(
        stat_df_ci,
        xmin = "group1",
        xmax = "group2",
        label = "label",
        y.position = "y.position",
        tip.length = 0.03,
        size = 14,
        bracket.size = 0.9
      )
    } else NULL
  } +
  labs(
    title = NULL,
    x = "Spatial Arrangement",
    y = "Travel Consistency\n(Net / Total Distance)"
  ) +
  theme_classic(base_size = 20) +
  theme(
    axis.line    = element_line(linewidth = 0.9, color = "black"),
    axis.text.y  = element_text(size = 18),
    axis.title.y = element_text(size = 20),
    axis.text.x  = element_text(size = 16, angle = 0, hjust = 0.5),
    axis.title.x = element_text(size = 16),
    legend.position = "none",
    strip.background = element_blank(),
    strip.text.x = element_text(size = 20),
    panel.spacing.x = unit(2, "lines"),
    plot.margin = margin(10, 10, 10, 10)
  )

ggsave(
  filename = "TravelConsistency_Seg_Clus.png",
  plot     = p_ci,
  width    = 7,
  height   = 4.7,
  dpi      = 600
)

p_ci
```


## 8.) Nearest Neighbor
```{r}
# MDAMB231 -> nearest Myoepithelial cell
library(RANN)

frame_interval_min <- 10
hours_per_frame <- frame_interval_min / 60

CP_clean_nn <- Combined_Positions %>%
  dplyr::mutate(
    frame_raw = suppressWarnings(readr::parse_number(as.character(Time))) %>% as.integer(),
    frame     = frame_raw - min(frame_raw, na.rm = TRUE), 
    Time_hr   = frame * hours_per_frame,
    X         = suppressWarnings(as.numeric(X_Position)),
    Y         = suppressWarnings(as.numeric(Y_Position))
  ) %>%
  dplyr::filter(
    is.finite(frame), is.finite(Time_hr), is.finite(X), is.finite(Y),
    Group %in% c("Segregated","Clustered"),
    CellTypeID %in% c("MDAMB231","Myoepithelial"),
    CellTypeID != "Missing_ID"
  )

compute_nn_mda_to_myo <- function(df_clean) {

  frames <- df_clean %>%
    dplyr::group_by(SampleID, Group, frame, Time_hr) %>%
    dplyr::group_split()

  purrr::map_dfr(frames, function(df_t) {

    sample_id <- df_t$SampleID[1]
    group     <- df_t$Group[1]
    fr        <- df_t$frame[1]
    thr       <- df_t$Time_hr[1]

    mda <- df_t %>% dplyr::filter(CellTypeID == "MDAMB231")
    myo <- df_t %>% dplyr::filter(CellTypeID == "Myoepithelial")

    if (nrow(mda) == 0 || nrow(myo) == 0) return(NULL)

    mda_mat <- as.matrix(mda[, c("X","Y")])
    myo_mat <- as.matrix(myo[, c("X","Y")])

    nn <- RANN::nn2(data = myo_mat, query = mda_mat, k = 1)

    idx <- nn$nn.idx[, 1]
    d1  <- nn$nn.dists[, 1]

    neighbor <- myo[idx, , drop = FALSE]

    tibble::tibble(
      SampleID       = sample_id,
      Group          = group,
      frame          = fr,
      Time_hr        = thr,
      MDA_UniqueID   = mda$UniqueID,
      MDA_X          = mda$X,
      MDA_Y          = mda$Y,
      NN_Dist_um     = d1,
      NN_UniqueID    = neighbor$UniqueID,
      NN_CellTypeID  = neighbor$CellTypeID
    )
  })
}

nn_mda_to_myo <- compute_nn_mda_to_myo(CP_clean_nn)

# quick sanity check
nn_mda_to_myo %>% dplyr::count(Group) %>% print()
nn_mda_to_myo %>% dplyr::summarise(min_d = min(NN_Dist_um, na.rm = TRUE),
                                  max_d = max(NN_Dist_um, na.rm = TRUE)) %>% print()

# summarize over time
nn_rep_time <- nn_mda_to_myo %>%
  dplyr::group_by(SampleID, Group, Time_hr) %>%
  dplyr::summarise(
    mean_dist_rep = mean(NN_Dist_um, na.rm = TRUE),
    n_cells       = sum(is.finite(NN_Dist_um)),
    .groups       = "drop"
  )

nn_time_summary <- nn_rep_time %>%
  dplyr::group_by(Group, Time_hr) %>%
  dplyr::summarise(
    mean_dist = mean(mean_dist_rep, na.rm = TRUE),
    sem_dist  = sd(mean_dist_rep, na.rm = TRUE) / sqrt(sum(is.finite(mean_dist_rep))),
    n_repl    = sum(is.finite(mean_dist_rep)),
    .groups   = "drop"
  ) %>%
  dplyr::mutate(Group = factor(Group, levels = c("Segregated","Clustered")))

# x-axis breaks (every 6 hours)
xmax_hr  <- max(nn_time_summary$Time_hr, na.rm = TRUE)
x_breaks <- seq(0, ceiling(xmax_hr / 6) * 6, by = 6)

p_nn_mean <- ggplot(
  nn_time_summary,
  aes(x = Time_hr, y = mean_dist, color = Group, fill = Group)
) +
  geom_ribbon(
    aes(ymin = pmax(mean_dist - sem_dist, 0),
        ymax = mean_dist + sem_dist),
    alpha = 0.18,
    color = NA
  ) +
  geom_line(linewidth = 1.0) +
  scale_color_manual(values = group_colors, drop = FALSE) +
  scale_fill_manual(values  = group_colors, drop = FALSE) +
  scale_x_continuous(breaks = x_breaks, expand = c(0, 0)) +
  labs(
    x = "Time (hours)",
    y = "TNBC to Nearest Myoepithelial \nNeighbor Mean Distance (µm)",
    title = NULL
  ) +
  theme_classic(base_size = 20) +
  theme(
    legend.position = "right",
    legend.title    = element_blank(),
    legend.text     = element_text(size = 16),
    axis.line       = element_line(linewidth = 0.9, color = "black"),
    axis.text       = element_text(size = 18),
    axis.title      = element_text(size = 20),
    strip.background = element_blank()
  )

ggsave(
  "NN_Distance_MDAMB231_to_Myoep_SegClus.png",
  p_nn_mean, width = 10, height = 4.8, dpi = 600
)

p_nn_mean
```



# xxxxxx
# COM for MyoEp
```{r}
frame_interval_min <- 10
hours_per_frame <- frame_interval_min / 60

group_cols <- group_colors  

myoep_id <- "Myoepithelial"

# 1) Clean positions
cp_clean_myoep <- Combined_Positions %>%
  dplyr::mutate(
    frame_raw = suppressWarnings(readr::parse_number(as.character(Time))) %>% as.integer(),
    X         = suppressWarnings(as.numeric(X_Position)),
    Y         = suppressWarnings(as.numeric(Y_Position))
  ) %>%
  dplyr::filter(
    CellTypeID == myoep_id,
    Group %in% c("Segregated", "Clustered"),
    is.finite(frame_raw), is.finite(X), is.finite(Y),
    !is.na(UniqueID)
  ) %>%
  dplyr::arrange(SampleID, Group, UniqueID, frame_raw) %>%
  dplyr::distinct(SampleID, UniqueID, frame_raw, .keep_all = TRUE) %>%
  dplyr::group_by(SampleID) %>%
  dplyr::mutate(
    frame0     = min(frame_raw, na.rm = TRUE),
    frame      = frame_raw - frame0,
    Time_hours = frame * frame_interval_min / 60
  ) %>%
  dplyr::ungroup() %>%
  dplyr::select(-frame0)

# 2) Per-sample COM at each frame
com_raw_myoep <- cp_clean_myoep %>%
  dplyr::group_by(SampleID, Group, frame) %>%
  dplyr::summarise(
    com_x      = mean(X, na.rm = TRUE),
    com_y      = mean(Y, na.rm = TRUE),
    Time_hours = dplyr::first(Time_hours),
    .groups    = "drop"
  )

# 3) Re-center per sample so t0 = (0,0)
com_centered_myoep <- com_raw_myoep %>%
  dplyr::group_by(SampleID, Group) %>%
  dplyr::arrange(Time_hours, .by_group = TRUE) %>%
  dplyr::mutate(
    com_x0 = dplyr::first(com_x[is.finite(com_x)]),
    com_y0 = dplyr::first(com_y[is.finite(com_y)]),
    dx     = com_x - com_x0,
    dy     = com_y - com_y0
  ) %>%
  dplyr::ungroup()

# 4) Mean ± SEM across replicates (SampleID) at each timepoint
com_summary_myoep <- com_centered_myoep %>%
  dplyr::group_by(Group, Time_hours) %>%
  dplyr::summarise(
    n_repl_dx = sum(is.finite(dx)),
    n_repl_dy = sum(is.finite(dy)),
    dx_mean   = mean(dx, na.rm = TRUE),
    dy_mean   = mean(dy, na.rm = TRUE),
    dx_sem    = dplyr::if_else(n_repl_dx >= 2, sd(dx, na.rm = TRUE) / sqrt(n_repl_dx), NA_real_),
    dy_sem    = dplyr::if_else(n_repl_dy >= 2, sd(dy, na.rm = TRUE) / sqrt(n_repl_dy), NA_real_),
    .groups   = "drop"
  ) %>%
  dplyr::mutate(Group = factor(Group, levels = c("Segregated", "Clustered")))

# 5) Helper to plot with fixed axis
plot_COM_axis <- function(df, axis_label) {
  ggplot(df, aes(x = Time_hours, y = mean, color = Group, fill = Group)) +
    geom_hline(yintercept = 0, linewidth = 0.6, color = "grey70") +
    geom_ribbon(aes(ymin = mean - sem, ymax = mean + sem), alpha = 0.18, color = NA) +
    geom_line(linewidth = 1.1, na.rm = TRUE) +
    scale_color_manual(values = group_cols, drop = FALSE) +
    scale_x_continuous(
      limits = c(0, 24),
      breaks = seq(0, 24, by = 6),
      expand = c(0, 0)
    ) +
    scale_fill_manual(values  = group_cols, drop = FALSE) +
    scale_y_continuous(
      limits = c(-200, 300),
      breaks = seq(-200, 300, by = 50)
    ) +
    labs(
      x = "Time (hours)",
      y = axis_label
    ) +
    theme_classic(base_size = 20) +
    theme(
      axis.line    = element_line(linewidth = 0.9, color = "black"),
      axis.text.y  = element_text(size = 18),
      axis.title.y = element_text(size = 16),
      axis.text.x  = element_text(size = 20, angle = 0, hjust = 0.5),
      axis.title.x = element_text(size = 20),
      legend.position = "right",
      legend.title = element_text(size = 18, margin = margin(b = 4)),
      legend.text = element_text(size = 16),
      legend.key.height = unit(10, "pt"),
      legend.spacing.y  = unit(2, "pt"),
      strip.background = element_blank(),
      strip.text.x = element_text(size = 20)
    )
}

# ΔX plot
dx_df_myoep <- com_summary_myoep %>%
  dplyr::select(Group, Time_hours, mean = dx_mean, sem = dx_sem)

P_COM_dx_Myoep <- plot_COM_axis(dx_df_myoep, axis_label = "ΔX Center of Mass Displacement (µm)")
ggsave("Myoepithelial_COM_Displacement_DX.png", P_COM_dx_Myoep, width = 7.5, height = 4.8, dpi = 600)

# ΔY plot
dy_df_myoep <- com_summary_myoep %>%
  dplyr::select(Group, Time_hours, mean = dy_mean, sem = dy_sem)

P_COM_dy_Myoep <- plot_COM_axis(dy_df_myoep, axis_label = "ΔY Center of Mass Displacement (µm)")
ggsave("Myoepithelial_COM_Displacement_DY.png", P_COM_dy_Myoep, width = 7.5, height = 4.8, dpi = 600)

P_COM_dx_Myoep
P_COM_dy_Myoep
```

```{r}
# time-average within each SampleID (replicate)
rep_avg_myoep <- com_centered_myoep %>%
  filter(is.finite(dx), is.finite(dy)) %>%
  group_by(SampleID, Group) %>%
  summarise(
    dX_mean = mean(dx, na.rm = TRUE),
    dY_mean = mean(dy, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(Group = factor(Group, levels = c("Segregated", "Clustered")))

df_box_myoep <- rep_avg_myoep %>%
  tidyr::pivot_longer(
    cols = c(dX_mean, dY_mean),
    names_to = "Component",
    values_to = "Value"
  ) %>%
  mutate(
    Component = recode(Component, dX_mean = "ΔX (µm)", dY_mean = "ΔY (µm)"),
    Component = factor(Component, levels = c("ΔX (µm)", "ΔY (µm)")),
    Group     = factor(Group, levels = c("Segregated", "Clustered"))
  )

# sanity
df_box_myoep %>% count(Component, Group)
summary(df_box_myoep$Value)
```

```{r}
star_from_p <- function(p) {
  if (is.na(p)) return(NA_character_)
  if (p <= 0.0001) return("****")
  if (p <= 0.001)  return("***")
  if (p <= 0.01)   return("**")
  if (p <= 0.05)   return("*")
  "ns"
}

welch_COM_by_component <- function(df_box,
                                  groups_levels = c("Segregated", "Clustered")) {

  df_box %>%
    group_by(Component) %>%
    group_modify(function(d, key) {

      d <- d %>%
        tidyr::drop_na(Value, Group) %>%
        mutate(Group = factor(Group, levels = groups_levels))

      # require both groups and >=2 replicates per group
      if (n_distinct(d$Group) < 2 || any(table(d$Group) < 2)) {
        return(tibble(
          comparison = "Segregated-Clustered",
          direction  = NA_character_,
          p_value    = NA_real_,
          sig        = NA_character_,
          statement  = paste0("Myoepithelial ", as.character(key$Component),
                              ": Not enough replicates per group")
        ))
      }

      tt <- t.test(Value ~ Group, data = d, var.equal = FALSE)
      means <- tapply(d$Value, d$Group, mean, na.rm = TRUE)

      dir_txt <- if (means[groups_levels[2]] > means[groups_levels[1]]) {
        paste0(groups_levels[2], " > ", groups_levels[1])
      } else {
        paste0(groups_levels[1], " > ", groups_levels[2])
      }

      p <- tt$p.value

      tibble(
        comparison = "Segregated-Clustered",
        direction  = dir_txt,
        p_value    = p,
        sig        = star_from_p(p),
        statement  = paste0(
          "Myoepithelial ", as.character(key$Component), ": ", dir_txt,
          " (Welch t-test, p = ", signif(p, 3), ", ", star_from_p(p), ")"
        )
      )
    }) %>%
    ungroup()
}

com_stats_myoep <- welch_COM_by_component(df_box_myoep)

com_stats_myoep
com_stats_myoep$statement
```

```{r}
fill_cols  <- group_colors
point_cols <- point_colors

sig_rows_myoep <- com_stats_myoep %>%
  filter(!is.na(sig), sig != "ns") %>%
  mutate(group1 = "Segregated", group2 = "Clustered")

# bracket y-positions per facet (data-driven, kept inside axis)
pval_df_myoep <- sig_rows_myoep %>%
  group_by(Component) %>%
  mutate(
    ymax_comp = max(df_box_myoep$Value[df_box_myoep$Component == Component[1]], na.rm = TRUE),

    bump = dplyr::case_when(
      Component == "ΔX (µm)" ~ 6,
      Component == "ΔY (µm)" ~ 14,
      TRUE                   ~ 8
    ),

    y.position = pmin(72, ymax_comp + bump),
    label = sig
  ) %>%
  ungroup() %>%
  select(Component, group1, group2, y.position, label)

p_timeavg_box_stars_myoep <- ggplot(df_box_myoep, aes(x = Group, y = Value, fill = Group)) +
  geom_hline(yintercept = 0, linewidth = 0.6, color = "grey50") +
  geom_boxplot(width = 0.6, alpha = 0.9, color = "black", outlier.shape = NA) +
  geom_point(
    aes(color = Group),
    position = position_jitter(width = 0.30, height = 0),
    size = 2.5, alpha = 0.9,
    show.legend = FALSE
  ) +
  facet_grid(~ Component, switch = "x") +
  scale_fill_manual(values = fill_cols, drop = FALSE) +
  scale_color_manual(values = point_cols, drop = FALSE) +
  scale_y_continuous(
    limits = c(-100, 200),
    breaks = seq(-100, 200, by = 50),
    expand = c(0, 0)
  ) +
  {
    if (nrow(pval_df_myoep) > 0) {
      ggpubr::stat_pvalue_manual(
        data       = pval_df_myoep,
        label      = "label",
        xmin       = "group1",
        xmax       = "group2",
        y.position = "y.position",
        tip.length = 0.02,
        size       = 12,
        bracket.size = 0.6
      )
    } else NULL
  } +
  labs(x = NULL, y = "Time-averaged COM displacement (µm)") +
  theme_classic(base_size = 20) +
  theme(
    legend.position  = "none",
    strip.placement  = "outside",
    strip.background = element_blank(),
    axis.line        = element_line(linewidth = 1, color = "black"),
    axis.text.y      = element_text(size = 16),
    axis.title.y     = element_text(size = 15),
    axis.text.x      = element_text(size = 16, angle = 0, hjust = 0.5),
    strip.text.x     = element_text(size = 16, margin = margin(b = 2)),
    panel.spacing.x = unit(2, "lines")
  )

ggsave(
  "Myoepithelial_COM_TimeAveraged_dX_dY_boxplot_stars.png",
  p_timeavg_box_stars_myoep,
  width = 7.2, height = 4.4, dpi = 600
)

p_timeavg_box_stars_myoep
```

```{r}
com_vectors_repl_myoep <- com_centered_myoep %>%
  dplyr::filter(is.finite(Time_hours), is.finite(dx), is.finite(dy)) %>%
  dplyr::group_by(Group, SampleID) %>%
  dplyr::summarise(
    Time_end = max(Time_hours, na.rm = TRUE),
    dx_end   = dx[which.max(Time_hours)],
    dy_end   = dy[which.max(Time_hours)],
    bearing_deg = atan2(dy_end, dx_end) * 180 / pi,
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    Group = factor(Group, levels = c("Segregated", "Clustered"))
  )

p_repl_vec_myoep <- ggplot(com_vectors_repl_myoep, aes(color = Group)) +
  geom_hline(yintercept = 0, linewidth = 0.4, color = "grey80") +
  geom_vline(xintercept = 0, linewidth = 0.4, color = "grey80") +
  geom_segment(
    aes(x = 0, y = 0, xend = dx_end, yend = dy_end),
    arrow = arrow(length = unit(2.2, "mm"), type = "closed"),
    linewidth = 0.9,
    alpha = 0.8
  ) +
  scale_color_manual(values = group_cols, drop = FALSE) +
  scale_x_continuous(
    limits = c(-200, 200),
    breaks = seq(-200, 200, by = 50)
  ) +
  scale_y_continuous(
    limits = c(-200, 200),
    breaks = seq(-200, 200, by = 50)
  ) +
  coord_fixed(expand = FALSE) +
  labs(
    x = "ΔX COM Displacement (µm)",
    y = "ΔY COM Displacement (µm)"
  ) +
  theme_classic(base_size = 14) +
  theme(
    legend.position = "right",
    legend.title    = element_blank(),
    legend.text     = element_text(size = 14),
    axis.text       = element_text(size = 14),
    axis.title      = element_text(size = 14),
    axis.line       = element_line(linewidth = 0.3, color = "black")
  )

ggsave("Myoepithelial_COM_EndVectors_ByReplicate.png",
       p_repl_vec_myoep, width = 6, height = 4, dpi = 600)

p_repl_vec_myoep
```

